### 2021-01-16 manuscript.Rmd
version for Taolin's thesis

---
output: word_document
bibliography: /home/xiang/Documents/Bibliography/seahorse.bib
---

Bayesian hierarchical modeling for Seahorse extracellular flux oxygen consumption rate data analysis

Xiang Zhang, 
Taolin Yuan, 
Keijer Jaap,
Vincent C. J. de Boer

Human and Animal Physiology, Wageningen University, De Elst 1, 6708 WD Wageningen, The Netherlands

# Abstract

Mitochondrial dysfunction is involved in many complex diseases. 
Efficient and accurate evaluation of mitochondrial functionality is crucial for understanding pathology as well as facilitating novel therapeutic developments. 
As a reliable platform, Seahorse extracellular flux (XF) analyzer is widely used for measuring mitochondrial oxygen consumption rate (OCR) in living cells. 
A hidden feature of Seahorse XF OCR data is that it has a complex data structure, caused by nesting and crossing between multiple factors. 
Surprisingly, statistical analysis of Seahorse XF data has not received sufficient attention, and current methods completely ignore the complex data structure, impairing the robustness of statistical inference. 
To rigorously incorporate the complex structure into data analysis, here we developed a Bayesian hierarchical modeling framework and demonstrated its applicability for analyzing Seahorse XF OCR data with two in-house experimental data sets. 
We showed that our approach can propagate uncertainty in the OCR measurements into the posterior estimation of bioenergetic measures, and is flexible to be applied to analyze experimental data with various experimental designs and sample sizes.

# Introduction

Mitochondria are double-membrane organelles that are central hubs in regulating energy generation and partitioning. 
Since defects in mitochondrial function have been found to be involved in cardiovascular diseases [@Ballinger2005], type II diabetes [@Szendroedi2011], neurodegenerative disease [@Johri2012] and cancer [@Erez2015], restoring mitochondrial function is emerging to be a therapeutic target for common diseases [@Murphy2018].

Mitochondria produce energy primarily by transferring electrons along the electron transport chain (ETC) on the inner membrane. 
Along the ETC, there are four complexes (including complex I, II, III and IV), building up a proton gradient that is ultimately used by ATP synthase to generate ATP. 
Since electrons are primarily accepted by O~2~ to produce H~2~O, the activity of ETC can be assessed by measuring oxygen consumption rate (OCR). 
As a reliable and efficient platform, Seahorse XF analyzer is a multiwell plate based respirometry assay that is widely used to quantify OCR in living cells [@Wu2007]. 
Typically, the seahorse XF analyzer measures the OCR of cells in a 96-well plate under different ETC perturbation scenarios, used for assessing mitochondrial functionality such as maximal respiration, leak respiration and ATP-linked respiration. 
Seahorse XF Cell Mito Stress Test consists of four phases including an initial phase, followed by three consecutive injections, including oligomycin, FCCP (carbonyl cyanide-4-(trifluoromethoxy)phenylhydrazone) and a mixture of rotenone and antimycin A. 
In the initial phase, no compound is injected and ETC is not perturbed. 
Once oligomycin is injected in the assay, proton translocation through ATP synthase is blocked, leading to a reduction in OCR. 
Next, injection of FCCP allows protons to move into the mitochondrial matrix independent of the ATP synthase, pushing the oxygen consumption to the maximum level. 
In the end, a mixture of rotenone and antimycin A is injected to inhibit complex I and III, shutting down mitochondrial respiration.

A standard Seahorse run includes three measurement cycles for each phase. 
Every measurement cycle starts by lowering the cartridge and creating a temporary semi-closed 7 $\mu$L chamber [@Gerencser2009]. 
During a measurement cycle, fluorescent oxygen sensors capture oxygen concentration changes in the chamber and output OCR. 
The Seahorse XF analyzer measures OCR in tens to hundreds of thousands of cells per well and typically requires 4–5 replicate wells per experimental group [@Artyomov2020].
A single 96-well plate can contain multiple experimental groups. 
Since measurements are nested within phases, phases are crossed with wells, and wells are nested in experimental groups, a very complex structure is inherently embedded in the Seahorse XF OCR data. 
Surprisingly, current Seahorse XF data analyses most often ignore this complex structure, and by default, data sets are often chopped into subgroups followed by performing ANOVA-like statistical tests. 
Other Seahorse XF data analysis tools such as OCR-Stats [@Yepez2018], SHORE [@Nicholas2017] and Bioenergetic Health Index [@Chacko2014] also do not take the complexity of the data structure into account.

A direct and rigorous way to incorporate the complex structure into data analysis is to use hierarchical modeling. 
Hierarchical models have been extensively developed for analyzing gene expression data  [@Robinson2010a;@Love2014a]. 
Here we developed a Bayesian hierarchical modeling framework for the Seahorse XF OCR data. 
Our modeling framework is based on a fully Bayesian approach due to its flexibility and convenience for constructing the hierarchical models. 
We used two in-house research cases to demonstrate the applicability of our approach analyzing Seahorse XF OCR data.

# Method

## Cell culture

Fibroblasts from controls and patients etc.

## Searhorse extracellular flux analysis

Cells were plated the day before the assay in 96 well plates etc…

## Cell number quantification

In both example studies, after Seahorse assays, cells in plate were stained with 4 $\mu$M of 2'-(4-ethoxyphenyl)-5-(4-methyl-1-piperazinyl)-2,5'-bi-1H-benzimidazole trihydrochloride trihydrate (Hoechst 33342, #B2261, Sigma-Aldrich) dissolved in assay medium, for 1 hr at 37°C non-CO 2 incubator. 
Images were taken by Cytation 1 Imaging Reader (BioTek) and cell number was determined by ImageJ 1.52.

## Case study 1

In the first study, fibroblast cells derived from a healthy subject were cultured in galactose exposure medium. 
A SIRT5 inhibitor, DK104e, or the control counterpart, DK104e[O] was administrated into the cell culture. 
Fibroblast treated with DMSO was used as the control. 
In this experiment, four dosages (10 $\mu$M, 20 $\mu$M, 60 $\mu$M and 70 $\mu$M) were used. 
Respiration of fibroblasts was measured using a Seahorse XFe96 Extracellular Flux Analyzer and a Mito Stress test was performed by following the manufacturer’s protocols.

During the quality control, we removed three wells (2 derived from 10 $\mu$M DK104e[O] group and 1 from DMSO group) due to negative OCR values. 
After quality control, the number of wells that contain cells treated by 10 $\mu$M, 20 $\mu$M, 60 $\mu$M and 70 $\mu$M DK104e were 10, 10, 10 and 9, respectively. 
Regarding DK104e[O], the number of wells that correspond to 10 $\mu$M, 20 $\mu$M, 60 $\mu$M and 70 $\mu$M were 8, 10, 9 and 9. 
There are 11 wells in which fibroblasts were treated with DMSO.

## Case study 2

In the second study, fibroblast cells derived from four healthy subjects and two patients with two homozygous loss-of-function variants in *SIRT5* were used for Seahorse XF analysis. 
One patient has a missense variant (c.340C>A, NM_012241.3) which leads to a proline114 substitution to threonine (P114T) in the SIRT5 protein. 
The other patient has missense mutation (c.382C>G, NM_012241.3) resulting in leucine128 to valine change (L128V). 
In this experiment, two injections were used. 
In particular, 1 $\mu$M oligomycin (#O4876, Sigma-Aldrich) and a mixture of 2.5 $\mu$M antimycin (#A8674, Sigma-Aldrich) and 1.25 $\mu$M rotenone (#R8875, Sigma-Aldrich).

For every fibroblast cell line derived from healthy controls, 8 replicate wells were used for galactose exposure and 7 replicate wells were used for glucose exposure. 
For each patient-derived fibroblast cell line, 8 replicate wells were used for galactose and glucose exposure, respectively.

## Implementation details

The Bayesian hierarchical models were implemented in Stan (version 2.19.3) [@Carpenter2017]. 
We fitted the model by running Hamiltonian Markov Chain Monte Carlo. 
We ran four Markov chains with 2000 iterations in each chain. 
Results were presented with the posterior mean with 95% credible interval. 
The code can be found at https://github.com/XiangZhangSC/seahorse

## Calculation of bioenergetic measures

Based on OCR (per 1000 cells), we calculated various bioenergetic measures, such as basal respiration, proton leak, ATP-linked respiration, spare respiratory capacity and maximal respiration.
These bioenergetic measures are defined as below.

$$\text{Basal respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, rotenone}$$
$$\text{ATP-linked respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, oligomycin}$$
$$\text{Proton leak} = \text{OCR}_\text{per 1k cells, oligomycin} - \text{OCR}_\text{per 1k cells, rotenone}$$
$$\text{Spare respiratory capacity} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, initial}$$
$$\text{Maximal respiration} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, rotenone}$$

# Results

## Bayesian hierarchical modeling framework for Seahorse OCR data analysis

In this study, we developed a Bayesian hierarchical modeling framework for the Seahorse XF OCR data. 
We applied a fully Bayesian approach due to its flexibility and convenience for constructing the hierarchical models. 
The mathematical description of our Bayesian hierarchical modeling framework is shown below:

$$ \text{OCR}_\text{obs}\text{[i]} \sim \text{Lognormal( logOCR}_\text{true}\text{[W[i],I[i]]}, \text{logOCR}_\text{se}\text{[W[i],I[i]])}$$

$$\text{exp(logOCR}_\text{true}\text{[W[i],I[i]])} = \text{OCR}_\text{per 1k cells}\text{[G[i], I[i]]} \times \text{N}_\text{cells}\text{[W[i]]}$$

$$\text{OCR}_\text{per 1k cells}\text{[G[i], I[i]]} \sim \text{Normal}(\mu\text{[i]}, \sigma\text{)} $$

$$\mu\text{[i]} = \text{Regression model depending on the experimental design}$$

$$\sigma \sim \text{Exponential(1)}$$

$$\text{logOCR}_\text{se}\text{[i]} \sim \text{prior distribution}$$

In Equation 6, we used a lognormal distribution with the true OCR value at the log scale ($\text{logOCR}_\text{true}$) and the measurement uncertainty (also at the log scale, $\text{logOCR}_\text{se}$) to model each observed OCR value ($\text{OCR}_\text{obs}$), because OCR values must be positive. 
$\text{OCR}_\text{obs}$ is a vector of length $\text{N}_\text{wells} \times \text{N}_\text{injections} \times \text{N}_\text{Measurement}$, where $\text{N}_\text{wells}$, $\text{N}_\text{injections}$ and $\text{N}_\text{Measurement}$ are number of wells, injections and measurement cycles, respectively. 
The $\text{logOCR}_\text{true}$ and $\text{logOCR}_\text{se}$ are matrices of $\text{N}_\text{wells}$ rows and $\text{N}_\text{injections}$ columns.

In Equation 7, we modeled the true OCR value as a function of the cell number. 
$\text{N}_\text{cells}$ is a vector of length $\text{N}_\text{wells}$, and every entry represents the cell number in that well. 
$\text{OCR}_\text{per 1k cells}$ is a matrix with $\text{N}_\text{Groups}$ rows and $\text{N}_\text{injections}$ columns. 
Every entry of this matrix represents OCR value per 1000 cells that received the same treatment (or from the same group) and injection.

In Equation 8 and 9, we applied a linear regression model in which the OCR value per 1000 cells was used as the response (or dependent) variable to estimate effect sizes of various factors (e.g. treatment, group). 
Since experimental design and plate set-up are not uniform between studies, regression models need to be constructed accordingly. 
The parameter $\sigma$ represented the residual that cannot be explained by the regression model. 
We used a generic the prior distribution ($\text{Exponential(1)}$) for $\sigma$.

In this study, since our case studies focused on human fibroblast cells, we used a informative prior distribution for $\text{logOCR}_\text{se}$. 
We estimated the prior distribution based on the Seahorse XF OCR data in OCR-stats [@Yepez2018] by running the fitdistr function built in the MASS r package. 
As a result, the $\text{Lognormal(-3.11, 0.81)}$ was used as the prior distribution for $\text{logOCR}_\text{se}$.

## Case studies

To demonstrate how our Bayesian hierarchical modeling approach can be used for Seahorse XF OCR data analysis, we used two in-house generated data sets that investigate mitochondrial function alteration due to either treatment of a novel inhibitor targeting SIRT5 (case 1) or a loss-of-function mutation in *SIRT5*.

### Case study 1: Factorial design with no biological replicates

In the first case study, we assessed if the SIRT5 inhibitor DK104e could lead to mitochondrial dysfunction in human fibroblasts.
This study used a factorial design with two factors: treatment and injection. 
In addition, a dosage escalation was applied in this study. 
Here, we treated the dosage factor as a continuous variable. 
Based on the experimental design of this study, we composed the regression model as:

$$\mu\text{[i]} = \beta_0 + \beta_T\text{[T[i]]} + \beta_D \text{D[i]} + \beta_I\text{[I[i]]}$$

$\beta_0$ is the overall average of OCR per 1000 cells in all wells and across injections. 
Based on the OCR-stats data, we estimated the prior distribution for $\beta_0$ as $\text{Lognormal(0.54, 0.32)}$. 
$\beta_T$ is the deviation in OCR due to treatment, DK104e (SIRT5 inhibitor) or DK104e[O] ("placebo"). 
$\beta_D$ is the effect when the dosage increases 10$\mu$M. 
We assume that the higher dosage the larger effect will be. 
$\beta_I$ is the deviation in OCR due to injection, including initial, oligomycin, FCCP and rotenone/antimycin A.

Our analysis showed that in DMSO-treated fibroblasts, the median basal respiration, ATP-linked respiration, proton leak, spare respiratory capacity and maximal respiration were 3.23 [95% credible interval 30.09, 3.34] pmol/min, 3.14 [2.99, 3.25] pmol/min, 0.0955 [0.0258, 0.158] pmol/min, -0.890 [-1.05, -0.705] pmol/min, and 2.34 [2.22, 2.48] pmol/min, respectively. 
In general, treating fibroblasts with 10-60 $\mu$M SIRT5 inhibitor Dk104e or "placebo" Dk104e[O] resulted in similar basal respiration, ATP-linked respiration, proton leak, spare respiratory capacity and maximal respiration. 
We found that the proton leak in 10 $\mu$M Dk104e treated fibroblasts was 0.114 [0.0143, 0.203] pmol/min, whereas in the "placebo" group (10 $\mu$M Dk104[O]) that value was 0.368 [0.101, 0.469] pmol/min. 
However, the difference was not statistically significant (mean difference -0.255 [-0.387, 0.0185]) (Figure 1). 
In contrast, we found that at 60 $\mu$M the proton leak in Dk104e treated fibroblasts was 0.144 [0.113, 0.181] pmol/min, 0.0736 [0.00898, 0.143] pmol/min higher than the proton leak (0.0711 [0.0124, 0.128]) in Dk104e[O] treated fibroblasts (Figure 1). 
At 70 $\mu$M, Dk104e treated fibroblasts showed increasing in basal respiration (1.45 [1.19, 1.68] pmol/min), ATP-linked respiration (0.793 [0.537, 1.08]), proton leak (0.662 [0.446, 0.784] pmol/min) and maximal respiration (1.09 [0.775, 1.40] pmol/min) (Figure 1).

![Figure 1 Posterior distributions of basal respiration, ATP-linked respiration,  proton leak, spare respiratory capacity and maximal respiration in galactose-cultured fibroblasts treated with either SIRT5 inhibitor Dk104e or Dk104e$\text{[O]}$](./inhibitor_study_respiration.jpg)

## Case study 2: Split-plot design

The second study was an experiment in which fibroblasts derived from two patients with *SIRT5* mutations were compared to fibroblasts derived from four healthy individuals. 
The goal of this study was to identify mitochondrial dysfunction in the *SIRT5* mutation carriers. 
In this study, subjects were nested in group factor (healthy control or patient). 
Meanwhile, treatment factor (galactose and glucose) and injection factor were crossed with subjects. 
With both nested and crossed factors, this experiment essentially has a split-plot design. 
The split plot design was originally invented in agricultural experiments in order to deal randomization of hard-to-change factors. 
Accordingly, we specified the regression model as:

$$\mu\text{[i]} = \beta_0 + 
\beta_G\text{[G[i]]} + 
\beta_T\text{[T[i]]} + 
\beta_I\text{[I[i]]} + 
\beta_{IT}\text{[I[i],T[i]]} + 
\beta_{GT}\text{[G[i],T[i]]} + 
\beta_{GI}\text{[G[i],I[i]]} + 
\beta_S\text{[S[i]]}$$

$\beta_0$ is the overall average OCR per 1000 cells in all wells and across injections. 
As before, we used $\text{Lognormal(0.54, 0.32)}$ as the prior distribution. 
$\beta_G$, $\beta_T$ and $\beta_I$ are OCR deflections due to group (healthy or patient), treatment (glucose or galactose), and injections, respectively. 
To take into account the fact that cell lines derived from the same subject were used for all levels of treatments and injections, $\beta_S$ is used to estimate the deflection due to subject factor. 
The interaction term, $\beta_{IT}$ denotes whether deflection due to injections is dependent on treatment factor. 
Similarly, the other two interaction terms, $\beta_{GT}$ and $\beta_{GI}$ denote whether deflection caused by treatment and injection factor are dependent on group factor.

Our analysis showed that on average 1000 fibroblasts treated with galactose had higher basal respiration (1.31 [0.801, 1.86] pmol/min) and ATP-linked respiration (1.17 [0.649, 1.71] pmol/min) than 1000 glucose-treated fibroblasts. 
In contrast, proton leak was similar in galactose or glucose-treated fibroblasts (median difference 0.149 [-0.371, 0.687]). 
Interestingly, our analysis showed that on average the basal respiration of galactose-treated fibroblasts (1000 cells) was 3.75 [3.33, 4.17] pmol/min in healthy controls and 4.60 [4.07, 5.13] pmol/min in patients with *SIRT5* mutations. 
In glucose-treated fibroblasts, the basal respiration of healthy controls was 2.42 [2.01, 2.84] pmol/min, whereas the basal respiration of patient-derived fibroblasts was 3.29 [2.75, 3.79] pmol/min. 
The basal respiration difference between patients and controls is 0.861 [0.293, 1.41] and was independent of treatment of glucose or galactose. 
The ATP-linked respiration in galactose-treated fibroblasts derived from healthy controls and patients with *SIRT5* mutations were 3.35 [2.94, 3.76] and 4.02 [3.47, 4.54], respectively. 
In glucose-treated fibroblasts, the ATP-linked respiration in healthy controls and patients were 2.17 [1.75, 2.59] pmol/min and 2.85 [2.29, 3.38] pmol/min, respectively. 
Similar to basal respiration, the difference in ATP-linked respiration between patients and controls was 0.682 [0.0998, 1.24] pmol/min, and this difference was independent of treatment of glucose or galactose. 
The proton leak of galactose-treated fibroblasts in healthy controls was 0.404 [-0.0152, 0.819] pmol/min whereas in patients the proton leak was 0.579 [0.0657, 1.12] pmol/min. 
In glucose-treated fibroblasts, healthy controls had proton leak 0.258 [-0.164, 0.665] pmol/min, and patients had proton leak 0.434 [-0.0804, 0.962] pmol/min. 
The proton leak difference between patients and controls was 0.176 [-0.363, 0.736] pmol/min and was independent of treatment of glucose or galactose.

![Figure 2 Posterior distributions of basal respiration, ATP-linked respiration and proton leak of galactose or glucose-treated fibroblasts derived from in healthy subjects and patients with loss-of-function mutations in *SIRT5*](./respiration_sirt5_mutation_study.jpg)

# Discussion

Although Seahorse XF analyzer is widely used in bioenergetic profiling, its data analysis has not received sufficient attention [@Yepez2018]. 
In the current study, we developed a Bayesian hierarchical modeling approach, which accounts for the complex structure of Seahorse OCR data. 
The complexity of Seahorse OCR data is due to the fact that Seahorse XF measurements are nested within injections, injections are crossed with wells, and wells are nested in experimental groups. 
The data structure can be even more complex when a study has a complicated experimental design with multiple factors.

From our perspective, there are three levels including 1) measurement, 2) well and 3) group in the Seahorse XF OCR data. 
At the lowest level, the measurement level, the variation within an injection phase are assumed to be caused by measurement error. 
This is because there are multiple measurement cycles during each injection phase, and every measurement cycle starts with a mix and wait steps to ensure the same baseline of cell values, cell physiology should not substantially change within an injection phase.
Since OCR data must be positive values, we used the lognormal distribution as the likelihood function. 
The same likelihood function was also used by OCR-Stats [@Yepez2018]. 
In the first example, we found that on average the uncertainty in FCCP-related measurements (0.147 [0.0443,0.270]) was larger than initial (0.0707 [0.0227,0.176]), oligomycin (0.0923 [0.0257,0.403]) and rotenone/antimycin A (0.0972 [0.0242,0.519]) (Figure S1).

![Figure S1 Uncertainty in OCR measurements per scenario in the example 1](./measurement_error_inhibitor_study.jpg)

The variable OCR values could indicate that FCCP possibly induces an unstable physiologic state, possibly due to the off-target effects. 
It has been proposed that other mitochondrial uncouplers such as BAM15 [@Kenwood2014] would have less off-target effects and would thus generate a more stable cellular physiologic state.

In the second example, we found that on average the measurement uncertainty in initial, oligomycin and rotenone/antimycin A were 0.0581 [0.0135, 0.221], 0.0760 [0.0220, 0.262] and 0.0798 [0.0208, 0.301], respectively. 
These measurement uncertainties were comparable between these two case studies, indicating that OCR values were stable in initial, oligomycin and rotenone/antimycin A scenario.

![Figure S2 Uncertainty in OCR measurements per scenario in the example 2](./measurement_error_mutation_study.jpg)

The second level of Seahorse OCR data is the well.  
Since the same cell lines with the same treatment were seeded in multiple wells, these replicated wells (also called pseudo replication) essentially reflect the same biological signals. 
In other words, these replicate wells should give the same OCR value if the number of cells are identical.
Treating these replicated wells as biological replicates in t test or ANOVA (analysis of variance) deviates from the assumptions that are associated with those procedures, making these statistical analysis not suitable for experimental data based on a single cell line. 
We notice that a "workaround" is to treat replicated experiments on different days or on different plates as a biological replication. 
However, this "workaround" does not solve the issue that there is no biological replicate. 
In fact, this issue is not unique to Seahorse OCR data analysis, it also happens to most if not all cell line-based studies. 
Our Bayesian hierarchical modeling approach provides a way to properly handle this type of experimental data. 
However, without biological replicates we still cannot generalize what we observed in one cell line to other cells. 
Furthermore, since the number of cells are not exactly the same in replicated wells, it is common to see OCR values differ among the pseudo replication [@Little2020]. 
To adjust for the cell number variation, our hierarchical model explicitly incorporated the cell number information of each wel0l (Equation 7) and estimated the $\text{OCR}_\text{per 1k cells}$, which represented the OCR value of 1000 cells.

The last level of Seahorse OCR data is the experimental group. 
$\text{OCR}_\text{per 1k cells}$ reflects OCR values that are determined by a particular cell line, a particular treatment and a particular injection. 
The cell line, treatment and their combination are viewed as groups. 
To find out the difference between groups, regression analysis of $\text{OCR}_\text{per 1k cells}$ needs to be performed.
In our case studies, we constructed two different regression models to handle Seahorse XF data with factorial and split-plot designs.

Our Bayesian hierarchical modeling approach has several advantages. 
The first advantage of our approach is that the uncertainty in the OCR measurements was incorporated in the posterior estimation of bioenergetic measures. 
The second advantage is that our approach is flexible and can be adapted to different experimental designs. 
The third advantage is that our approach can work with any sample size and unbalanced group sizes. 
For example, case study 2 has four biological replicates in the control group whereas two in the patient group. 
This is especially useful for the study that is challenging to acquire biological samples.

# Reference

### 2021-02-12 manuscript.Rmd
initial version of OCRbayes

@@
 bibliography: /home/xiang/Documents/Bibliography/seahorse.bib
+csl: plos-one.csl
 ---
 
-Bayesian hierarchical modeling for Seahorse extracellular flux oxygen consumption rate data analysis
+OCRbayes: A Bayesian hierarchical modeling framework for Seahorse extracellular flux oxygen consumption rate data analysis
 
@@
 Efficient and accurate evaluation of mitochondrial functionality is crucial for understanding pathology as well as facilitating novel therapeutic developments. 
-As a reliable platform, Seahorse extracellular flux (XF) analyzer is widely used for measuring mitochondrial oxygen consumption rate (OCR) in living cells. 
-A hidden feature of Seahorse XF OCR data is that it has a complex data structure, caused by nesting and crossing between multiple factors. 
+As a popular platform, Seahorse extracellular flux (XF) analyzer is widely used for measuring mitochondrial oxygen consumption rate (OCR) in living cells. 
+A hidden feature of Seahorse XF OCR data is that it has a complex data structure, caused by nesting and crossing between measurement cycles, wells and plates. 
 Surprisingly, statistical analysis of Seahorse XF data has not received sufficient attention, and current methods completely ignore the complex data structure, impairing the robustness of statistical inference.
-To rigorously incorporate the complex structure into data analysis, here we developed a Bayesian hierarchical modeling framework and demonstrated its applicability for analyzing Seahorse XF OCR data with two in-house experimental data sets.
-We showed that our approach can propagate uncertainty in the OCR measurements into the posterior estimation of bioenergetic measures, and is flexible to be applied to analyze experimental data with various experimental designs and sample sizes.
+To rigorously incorporate the complex structure into data analysis, here we developed a Bayesian hierarchical modeling framework, OCRbayes, and demonstrated its applicability based on analysis of published data sets.
+We showed that OCRbayes can analyze seahorse XF OCR data with either single or multiple plates, and it has potential to be used for diagnosing patients with mitochondrial diseases.
 
@@
 As a reliable and efficient platform, Seahorse XF analyzer is a multiwell plate based respirometry assay that is widely used to quantify OCR in living cells [@Wu2007]. 
-Typically, the seahorse XF analyzer measures the OCR of cells in a 96-well plate under different ETC perturbation scenarios, used for assessing mitochondrial functionality such as maximal respiration, leak respiration and ATP-linked respiration.
+Typically, the seahorse XF analyzer measures the OCR of cells in a 96-well plate under different ETC perturbation scenarios, used for assessing mitochondrial functionality such as maximal respiration, leak respiration and ATP-linked respiration [@@Chacko2014].
 Seahorse XF Cell Mito Stress Test consists of four phases including an initial phase, followed by three consecutive injections, including oligomycin, FCCP (carbonyl cyanide-4-(trifluoromethoxy)phenylhydrazone) and a mixture of rotenone and antimycin A.
@@
 
-A standard Seahorse run includes three measurement cycles for each phase. 
+A standard seahorse run includes three measurement cycles for each phase. 
 Every measurement cycle starts by lowering the cartridge and creating a temporary semi-closed 7 $\mu$L chamber [@Gerencser2009]. 
-During a measurement cycle, fluorescent oxygen sensors capture oxygen concentration changes in the chamber and output OCR. 
+During a measurement cycle, fluorescent oxygen sensors capture oxygen concentration changes in the chamber and outputs OCR. 
 The Seahorse XF analyzer measures OCR in tens to hundreds of thousands of cells per well and typically requires 4–5 replicate wells per experimental group [@Artyomov2020].
 A single 96-well plate can contain multiple experimental groups. 
-Since measurements are nested within phases, phases are crossed with wells, and wells are nested in experimental groups, a very complex structure is inherently embedded in the Seahorse XF OCR data. 
+Since measurement cycles are nested within phases, phases are crossed with wells, and wells are nested in experimental groups, a very complex structure is inherently embedded in the Seahorse XF OCR data.
 Surprisingly, current Seahorse XF data analyses most often ignore this complex structure, and by default, data sets are often chopped into subgroups followed by performing ANOVA-like statistical tests.
-Other Seahorse XF data analysis tools such as OCR-Stats [@Yepez2018], SHORE [@Nicholas2017] and Bioenergetic Health Index [@Chacko2014] also do not take the complexity of the data structure into account.
+Although there are tools developed for seahorse data analysis, none of them takes the complexity of the data structure into account. 
+In addition, these methods differ in how to extract OCR value for each phase. 
+For instance, OCR-Stats [@Yepez2018] took OCR value from the last measurement cycle in phase 1, the minimal OCR value in phase 2, the maximal OCR value in phase 3 and the minimal OCR value in phase 4.
+In contrast, SHORE [@Nicholas2017] took the mean OCR value in phase 1, the median OCR value in phase 2, the maximal OCR value in phase 3, and the median OCR value in phase 4.
 
-A direct and rigorous way to incorporate the complex structure into data analysis is to use hierarchical modeling. 
-Hierarchical models have been extensively developed for analyzing gene expression data  [@Robinson2010a;@Love2014a]. 
-Here we developed a Bayesian hierarchical modeling framework for the Seahorse XF OCR data. 
-Our modeling framework is based on a fully Bayesian approach due to its flexibility and convenience for constructing the hierarchical models. 
-We used two in-house research cases to demonstrate the applicability of our approach analyzing Seahorse XF OCR data.
+A natural way to incorporate the complex structure into data analysis is to use hierarchical modeling, which has been extensively developed for analyzing gene expression data  [@Robinson2010a;@Love2014a].
+Here we developed a Bayesian hierarchical modeling framework, OCRbayes, for the Seahorse XF OCR data analysis. 
+OCRbayes is based on a fully Bayesian approach due to its flexibility and convenience for constructing the hierarchical models. 
+We analyzed a published OCR data set OCR data set to demonstrate the applicability of our approach analyzing Seahorse XF OCR data as well as its potential usage for mitochondrial disease diagnostics.
 
@@
 
-## Cell culture
+To demonstrate how OCRbayes can be used for Seahorse XF OCR data analysis, we analyzed 26 human fibroblasts that were derived from patients that were suspected to have mitochondrial dysfunction. 
+In particular, six of these cell lines were derived from patients with rare mutations in genes associated with an established cellular respiratory defect, and they serve as positive controls. 
+Meanwhile, there are two cell lines that showed no respiratory defects in earlier studies [@Powell2015;@Kremer2016] and they served as negative controls. 
+Since fold change of maximal respiration is the most common metric used for identifying mitochondrial dysfunction, we compared the maximal respiration of all these 26 cell lines against the control cell line.
 
-Fibroblasts from controls and patients etc.
+## Data
 
-## Searhorse extracellular flux analysis
+We downloaded the seahorse XF OCR data presented in OCR-Stats [@Yepez2018] from the journal website https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0199938#sec022. 
+This data set contains 203 human fibroblasts assayed in 126 plates. 
+The normal human dermal fibroblasts (NHDF) is used as control. 
+The rest 202 cell lines were derived from patients suspected to suffer from rare mitochondrial diseases. 
+Among these 202 cell lines, 26 fibroblast cell lines were grown independently and were measured in multiple plates. 
+The rest 176 fibroblast cell lines were measured in a single plate. 
+We then spitted the data set into two subsets. 
+One subset, which contains the 176 fibroblasts that were assayed in a single plate, was used for estimating prior distributions. 
+The other subset, which contains the 26 fibroblasts, was used for benchmarking.
 
-Cells were plated the day before the assay in 96 well plates etc…
-
-## Cell number quantification
-
-In both example studies, after Seahorse assays, cells in plate were stained with 4 $\mu$M of 2'-(4-ethoxyphenyl)-5-(4-methyl-1-piperazinyl)-2,5'-bi-1H-benzimidazole trihydrochloride trihydrate (Hoechst 33342, #B2261, Sigma-Aldrich) dissolved in assay medium, for 1 hr at 37°C non-CO 2 incubator.
-Images were taken by Cytation 1 Imaging Reader (BioTek) and cell number was determined by ImageJ 1.52.
-
-## Case study 1
-
-In the first study, fibroblast cells derived from a healthy subject were cultured in galactose exposure medium. 
-A SIRT5 inhibitor, DK104e, or the control counterpart, DK104e[O] was administrated into the cell culture. 
-Fibroblast treated with DMSO was used as the control. 
-In this experiment, four dosages (10 $\mu$M, 20 $\mu$M, 60 $\mu$M and 70 $\mu$M) were used. 
-Respiration of fibroblasts was measured using a Seahorse XFe96 Extracellular Flux Analyzer and a Mito Stress test was performed by following the manufacturer’s protocols.
-
-During the quality control, we removed three wells (2 derived from 10 $\mu$M DK104e[O] group and 1 from DMSO group) due to negative OCR values. 
-After quality control, the number of wells that contain cells treated by 10 $\mu$M, 20 $\mu$M, 60 $\mu$M and 70 $\mu$M DK104e were 10, 10, 10 and 9, respectively. 
-Regarding DK104e[O], the number of wells that correspond to 10 $\mu$M, 20 $\mu$M, 60 $\mu$M and 70 $\mu$M were 8, 10, 9 and 9. 
-There are 11 wells in which fibroblasts were treated with DMSO.
-
-## Case study 2
-
-In the second study, fibroblast cells derived from four healthy subjects and two patients with two homozygous loss-of-function variants in *SIRT5* were used for Seahorse XF analysis. 
-One patient has a missense variant (c.340C>A, NM_012241.3) which leads to a proline114 substitution to threonine (P114T) in the SIRT5 protein. 
-The other patient has missense mutation (c.382C>G, NM_012241.3) resulting in leucine128 to valine change (L128V). 
-In this experiment, two injections were used. 
-In particular, 1 $\mu$M oligomycin (#O4876, Sigma-Aldrich) and a mixture of 2.5 $\mu$M antimycin (#A8674, Sigma-Aldrich) and 1.25 $\mu$M rotenone (#R8875, Sigma-Aldrich).
-
-For every fibroblast cell line derived from healthy controls, 8 replicate wells were used for galactose exposure and 7 replicate wells were used for glucose exposure. 
-For each patient-derived fibroblast cell line, 8 replicate wells were used for galactose and glucose exposure, respectively.
+Since dealing missing observations is not the focus of this study, we removed the entire measurement values of a well if it contains a missing OCR value. 
+In addition, since we focused on the standard mitochondrial stress test assay, we removed any well if that well does not have 4 phases with 3 measurement cycles each. 
+For the prior distribution subset, we had 177 (including the control cell line NHDF) cell lines on 78 plates after filtering. 
+Since NHDF cell line was present in all 78 plates, the OCR values derived from NHDF cell line were used for estimating the prior distribution of between plate variation.
 
@@
 We ran four Markov chains with 2000 iterations in each chain. 
-Results were presented with the posterior mean with 95% credible interval. 
-The code can be found at https://github.com/XiangZhangSC/seahorse
+The code can be found at https://github.com/XiangZhangSC/seahorse. 
+We also provided the R code of OCRbayes in the supplementary file.
 
@@
 
-$$\text{Basal respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, rotenone}$$
-$$\text{ATP-linked respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, oligomycin}$$
-$$\text{Proton leak} = \text{OCR}_\text{per 1k cells, oligomycin} - \text{OCR}_\text{per 1k cells, rotenone}$$
-$$\text{Spare respiratory capacity} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, initial}$$
-$$\text{Maximal respiration} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, rotenone}$$
+$$\text{Basal respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, rotenone} \text{ (Equation 1)}$$
+$$\text{ATP-linked respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, oligomycin} \text{ (Equation 2)}$$
+$$\text{Proton leak} = \text{OCR}_\text{per 1k cells, oligomycin} - \text{OCR}_\text{per 1k cells, rotenone} \text{ (Equation 3)}$$
+$$\text{Spare respiratory capacity} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, initial} \text{ (Equation 4)}$$
+$$\text{Maximal respiration} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, rotenone} \text{ (Equation 5)}$$
+
+## Statistical analysis
+
+For each cell line study, we compared the patient cell line to the control cell line (NHDF). 
+Since maximal respiration is the primary readout of mitochondrial defect in literature, and was also evaluated in the OCR-Stats, here we reported the mean log2 fold change of maximal respiration together with the False Discovery Rate (FDR).
+FDR was calculated based on the posterior error probability ($1 - P(log2(\frac{\text{patient}}{\text{control}}|\text{data}) < 0)$), where $P$ represents the posterior distribution after fitting the Bayesian model to the experimental OCR data.
+If the FDR below 0.05, we consider that difference between patient and control cell line was statistically significant.
 
@@
 
-## Bayesian hierarchical modeling framework for Seahorse OCR data analysis
+## OCRbayes: A Bayesian hierarchical modeling framework for Seahorse OCR data analysis
 
-In this study, we developed a Bayesian hierarchical modeling framework for the Seahorse XF OCR data. 
-We applied a fully Bayesian approach due to its flexibility and convenience for constructing the hierarchical models. 
+In order to incorporate the complex data structure into the analysis of seahorse XF OCR measurements, we developed a Bayesian hierarchical modeling framework, OCRbayes. 
+In this study, we focused on experimental data containing two groups, patient and control. 
 The mathematical description of our Bayesian hierarchical modeling framework is shown below:
 
-$$ \text{OCR}_\text{obs}\text{[i]} \sim \text{Lognormal( logOCR}_\text{true}\text{[W[i],I[i]]}, \text{logOCR}_\text{se}\text{[W[i],I[i]])}$$
+$$\text{OCR}_\text{obs}\text{[i]} \sim \text{Lognormal(logOCR}_\text{true}\text{[P[i],W[i],I[i]]}, \text{logOCR}_\text{se}\text{[P[i],I[i]])} \text{ (Equation 6)}$$
+$$\text{logOCR}_\text{true}\text{[P[i],W[i],I[i]]} \sim \text{Normal(}\text{OCR}_\text{per 1k cells}\text{[P[i],G[i],I[i]]} \times \text{N}_\text{cells}\text{[W[i]]}, \sigma_\text{well}\text{[P[i],G[i],I[i]])} \text{ (Equation 7)}$$
+$$\text{OCR}_\text{per 1k cells}\text{[P[i],G[i],I[i]]} \sim \text{Lognormal(}\mu_{\text{OCR}_\text{per 1k cells}}\text{[G[i], I[i]]}, \sigma_\text{plate}\text{[G[i], I[i]])} \text{ (Equation 8)}$$
 
-$$\text{exp(logOCR}_\text{true}\text{[W[i],I[i]])} = \text{OCR}_\text{per 1k cells}\text{[G[i], I[i]]} \times \text{N}_\text{cells}\text{[W[i]]}$$
+From our perspective, there are three sources of technical variations including 1) between measurement cycle variation, 2) between well variation and 3) between plate variation in the Seahorse XF OCR data.
 
-$$\text{OCR}_\text{per 1k cells}\text{[G[i], I[i]]} \sim \text{Normal}(\mu\text{[i]}, \sigma\text{)} $$
+The between measurement cycle variation refers to the variability between the three OCR values generated by the three measurement cycles within an injection phase. 
+In Equation 6, we used a lognormal distribution with the true OCR value at the log scale ($\text{logOCR}_\text{true}$) and the between measurement cycle standard deviation (also at the log scale, $\text{logOCR}_\text{se}$) to model each observed OCR value ($\text{OCR}_\text{obs}$).
+$\text{OCR}_\text{obs}$ is a vector of length $\text{N}_\text{plates} \times \text{N}_\text{wells} \times \text{N}_\text{injections} \times \text{N}_\text{Measurement}$, where $\text{N}_\text{plates}$, $\text{N}_\text{wells}$, $\text{N}_\text{injections}$ and $\text{N}_\text{Measurement}$ are number of plates, wells, injections and measurement cycles, respectively.
+The $\text{logOCR}_\text{true}$ refers to $\text{N}_\text{plates}$ layers of matrices. 
+Each layer is a matrix with $\text{N}_\text{wells}$ rows and $\text{N}_\text{injections}$ columns. 
+$\text{logOCR}_\text{se}$ is a matrix of $\text{N}_\text{plates}$ rows and $\text{N}_\text{injections}$ columns.
 
-$$\mu\text{[i]} = \text{Regression model depending on the experimental design}$$
+In a typical seahorse XF assay, the same cell line from the same experimental group is often seeded in multiple wells. 
+However the number of cells in these replicate wells are not identical, thus causing variation in OCR values between these replicate wells. 
+To adjust for the effect of cell number difference, in Equation 7 we modeled the true OCR value as a function of the cell number, and from there estimated $\text{OCR}_\text{per 1k cells}$, which represents OCR value per 1000 cells that received the same treatment (or from the same group) and injection.
+$\text{OCR}_\text{per 1k cells}$ is a three dimensional matrix with $\text{N}_\text{Groups}$ layers, and each layer has $\text{N}_\text{Groups}$ rows and $\text{N}_\text{injections}$ columns. 
+$\text{N}_\text{cells}$ is a vector of length $\text{N}_\text{wells}$, and every entry represents the cell number in that well. 
+In addition to cell number difference, initial conditions, treatment concentration, or fluorophore sleeve calibration can also contribute to between well variation. 
+Since these factors were difficult to be measured experimentally, we can only estimate an overall effect based on the residual parameter $\sigma_\text{well}$. 
+Thus, $\sigma_\text{well}$ represents the well-to-well variation after accounting for cell number difference.
 
-$$\sigma \sim \text{Exponential(1)}$$
+Last but not the least, the seahorse XF assay for a patient cell line is often repeated on different days and on more than one plates. 
+Due to differences in temperature, seeding time, growth time, growth medium, sensor cartridge as well as treatment efficiency, the OCR outcomes will differ between plates. 
+To take into account the between plate variation, in Equation 8 we used another lognormal distribution with the true OCR value per 1000 cells at log scale ($\mu_{\text{OCR}_\text{per 1k cells}}$) and the between plate standard deviation ($\sigma_\text{plate}$) to model the OCR value per 1000 cells $\text{OCR}_\text{per 1k cells}$.
 
-$$\text{logOCR}_\text{se}\text{[i]} \sim \text{prior distribution}$$
+In this study, since our case studies focused on human fibroblast cells, we used informative prior distributions for $\text{logOCR}_\text{se}$, $\sigma_\text{well}$, $\sigma_\text{plate}$ and $\mu_{\text{OCR}_\text{per 1k cells}}$.
+Since values of all these parameters must be positive, we used four lognormal distributions. 
+We estimated the lognormal distribution parameters based on the Seahorse XF OCR data in OCR-stats [@Yepez2018] by running the fitdistr function built in the MASS r package. 
+In this study, our prior distributions are
+$$
+\begin{align}
+\text{logOCR}_\text{se} & \sim \text{Lognormal(-3.23, 0.79)} \\
+\sigma_\text{well} & \sim \text{Lognormal(-1.62, 0.53)} \\
+\sigma_\text{plate} & \sim \text{Lognormal(-1.18, 0.05)} \\
+\mu_{\text{OCR}_\text{per 1k cells}} & \sim \text{Lognormal(0.3, 0.79)}
+\end{align}
+$$
 
-In Equation 6, we used a lognormal distribution with the true OCR value at the log scale ($\text{logOCR}_\text{true}$) and the measurement uncertainty (also at the log scale, $\text{logOCR}_\text{se}$) to model each observed OCR value ($\text{OCR}_\text{obs}$), because OCR values must be positive.
-$\text{OCR}_\text{obs}$ is a vector of length $\text{N}_\text{wells} \times \text{N}_\text{injections} \times \text{N}_\text{Measurement}$, where $\text{N}_\text{wells}$, $\text{N}_\text{injections}$ and $\text{N}_\text{Measurement}$ are number of wells, injections and measurement cycles, respectively.
-The $\text{logOCR}_\text{true}$ and $\text{logOCR}_\text{se}$ are matrices of $\text{N}_\text{wells}$ rows and $\text{N}_\text{injections}$ columns.
+## OCRbayes uses OCR measurements as inputs and outputs posterior distributions for respiration metrics
 
-In Equation 7, we modeled the true OCR value as a function of the cell number. 
-$\text{N}_\text{cells}$ is a vector of length $\text{N}_\text{wells}$, and every entry represents the cell number in that well. 
-$\text{OCR}_\text{per 1k cells}$ is a matrix with $\text{N}_\text{Groups}$ rows and $\text{N}_\text{injections}$ columns. 
-Every entry of this matrix represents OCR value per 1000 cells that received the same treatment (or from the same group) and injection.
+OCRbayes works with the raw OCR data, and estimates the OCR per 1000 cells during initial phase (Int1), oligomycin phase (Int2), FCCP phase (Int3) and rotenone phase (Int4) (Figure 1A and 1B). 
+Afterwards, the posterior OCR per 1000 cells values were transformed into various respiration metrics, such as ATP-linked respiration, basal respiration, maximal respiration, proton leak and spare respiratory capacity (Figure 1C).
+As an example, here we showed patient 65126 and the control cell line (NHDF) on the corresponding plates. 
+The posterior distributions showed the range of values for a respiration parameter. 
+The spread of the posterior distributions represented the uncertainty, which is resulted from technical variations that will be discussed in the next section.
 
-In Equation 8 and 9, we applied a linear regression model in which the OCR value per 1000 cells was used as the response (or dependent) variable to estimate effect sizes of various factors (e.g.
-treatment, group).
-Since experimental design and plate set-up are not uniform between studies, regression models need to be constructed accordingly. 
-The parameter $\sigma$ represented the residual that cannot be explained by the regression model. 
-We used a generic the prior distribution ($\text{Exponential(1)}$) for $\sigma$.
+![Figure 1 (A) Raw OCR values for patient 65126 and control cell line in two plates.
+(B) Posterior distributions for OCR per 1000 cells in patient 65126 and control cell line during initial phase (Int1), oligomycin phase (Int2), FCCP phase (Int3) and rotenone phase (Int4).
+(C) Posterior distributions for respiration metrics in patient 65126 and control cell line.](./posterior_respiration_examples.jpeg)
 
-In this study, since our case studies focused on human fibroblast cells, we used a informative prior distribution for $\text{logOCR}_\text{se}$. 
-We estimated the prior distribution based on the Seahorse XF OCR data in OCR-stats [@Yepez2018] by running the fitdistr function built in the MASS r package. 
-As a result, the $\text{Lognormal(-3.11, 0.81)}$ was used as the prior distribution for $\text{logOCR}_\text{se}$.
+## OCRbayes accounts for various technical variations during seahorse XF OCR data analysis
 
-## Case studies
+To visualize the technical variations, we used the cell line 76065 as an example. 
+This cell line was measured on five different plates, allowing us to visualize all three technical variations.  
+We observed that in the initial phase (Int1) and FCCP phase (Int3), the between plate variation is dominant source of technical variation, followed by between well and between measurement cycle variations.
+However, in the oligomycin phase (Int2) and rotenone phase (Int4), the between well variation can be as important as the between plate variation.
 
-To demonstrate how our Bayesian hierarchical modeling approach can be used for Seahorse XF OCR data analysis, we used two in-house generated data sets that investigate mitochondrial function alteration due to either treatment of a novel inhibitor targeting SIRT5 (case 1) or a loss-of-function mutation in *SIRT5*.
+![Figure 2 Technical variations in seahorse XF OCR data of experiments based on cell line 76065 and control cell line.
+Int1, Int2, Int3 and Int4 represent initial, oligomycin, FCCP and rotenone phase.
+All the variations were on log scale.](./technical_variations_76065.jpeg)
 
-### Case study 1: Factorial design with no biological replicates
+## Benchmark
 
-In the first case study, we assessed if the SIRT5 inhibitor DK104e could lead to mitochondrial dysfunction in human fibroblasts.
-This study used a factorial design with two factors: treatment and injection. 
-In addition, a dosage escalation was applied in this study. 
-Here, we treated the dosage factor as a continuous variable. 
-Based on the experimental design of this study, we composed the regression model as:
+To demonstrate that OCRbayes works properly, we applied it to analyze a published seahorse XF OCR data set containing 26 patient cell lines as well as a control cell line. 
+In a previous study, 6 cell lines were labeled as positive controls and 2 cell lines were labeled as negative controls. 
+Based on our analysis, we found six patient cell lines that had lower maximal respiration compared to the control cell line with False Discovery Rate (FDR) below 0.05. 
+Five of these six cell lines were actually the positive controls. 
+Meanwhile, the two negative controls had FDR above 0.05.
 
-$$\mu\text{[i]} = \beta_0 + \beta_T\text{[T[i]]} + \beta_D \text{D[i]} + \beta_I\text{[I[i]]}$$
+![Figure 3 Maximal respiration change patient vs.
+control on multiple plates.
+(A) average log2 fold change (y-axis) of maximal respiration of all cell lines repeated across plates (x-axis) and their respective controls, sorted by the False Discovery Rate (FDR).
+Red and blue dots represent positive and negative controls, respectively.
+(B) similar to (A), but depicting FDR.
+Red dashed line represents FDR = 0.05.](./benchmark.jpeg)
 
-$\beta_0$ is the overall average of OCR per 1000 cells in all wells and across injections. 
-Based on the OCR-stats data, we estimated the prior distribution for $\beta_0$ as $\text{Lognormal(0.54, 0.32)}$. 
-$\beta_T$ is the deviation in OCR due to treatment, DK104e (SIRT5 inhibitor) or DK104e[O] ("placebo"). 
-$\beta_D$ is the effect when the dosage increases 10$\mu$M. 
-We assume that the higher dosage the larger effect will be. 
-$\beta_I$ is the deviation in OCR due to injection, including initial, oligomycin, FCCP and rotenone/antimycin A.
+Our analysis successfully recalled  all negative controls and five out of six positive controls. 
+Interestingly, our analysis showed that there was no significant reduction of maximal respiration in patient 61818, which was used as a positive control. 
+This patient cell line was measured by seahorse XF OCR assay on three independent plates on three different days. 
+We analyzed these three repeated experiments separately as well as combined them together. 
+We found that only the second experiment showed difference in maximal respiration.
 
-Our analysis showed that in DMSO-treated fibroblasts, the median basal respiration, ATP-linked respiration, proton leak, spare respiratory capacity and maximal respiration were 3.23 [95% credible interval 30.09, 3.34] pmol/min, 3.14 [2.99, 3.25] pmol/min, 0.0955 [0.0258, 0.158] pmol/min, -0.890 [-1.05, -0.705] pmol/min, and 2.34 [2.22, 2.48] pmol/min, respectively.
-In general, treating fibroblasts with 10-60 $\mu$M SIRT5 inhibitor Dk104e or "placebo" Dk104e[O] resulted in similar basal respiration, ATP-linked respiration, proton leak, spare respiratory capacity and maximal respiration.
-We found that the proton leak in 10 $\mu$M Dk104e treated fibroblasts was 0.114 [0.0143, 0.203] pmol/min, whereas in the "placebo" group (10 $\mu$M Dk104[O]) that value was 0.368 [0.101, 0.469] pmol/min.
-However, the difference was not statistically significant (mean difference -0.255 [-0.387, 0.0185]) (Figure 1). 
-In contrast, we found that at 60 $\mu$M the proton leak in Dk104e treated fibroblasts was 0.144 [0.113, 0.181] pmol/min, 0.0736 [0.00898, 0.143] pmol/min higher than the proton leak (0.0711 [0.0124, 0.128]) in Dk104e[O] treated fibroblasts (Figure 1).
-At 70 $\mu$M, Dk104e treated fibroblasts showed increasing in basal respiration (1.45 [1.19, 1.68] pmol/min), ATP-linked respiration (0.793 [0.537, 1.08]), proton leak (0.662 [0.446, 0.784] pmol/min) and maximal respiration (1.09 [0.775, 1.40] pmol/min) (Figure 1).
+![Figure 4 (A) Raw OCR values for patient cell line 61818 in three repeated plates.
+(B) Posterior distributions of maximal respiration of patient 61818 and the control cell line in the same plate.
+The top three panels represent analysis of three repeated experiments separately.
+The last panel represents the data analysis outcome after combing all three repeated experiments.](./posterior_maximal_respiration_61818.jpeg)
 
-![Figure 1 Posterior distributions of basal respiration, ATP-linked respiration,  proton leak, spare respiratory capacity and maximal respiration in galactose-cultured fibroblasts treated with either SIRT5 inhibitor Dk104e or Dk104e$\text{[O]}$](./inhibitor_study_respiration.jpg)
+## OCRbayes can facilitate decision making about whether testing on multiple plates is necessary
 
-## Case study 2: Split-plot design
+A frequent question regarding seahorse XF analysis is that whether it is enough to perform the assay on a single plate or do we really need repeat the assay multiple times on different plates. 
+To address this question, we used the two positive control cell lines 67375 and 76065 as examples, since in our analysis they were shown to have the maximal and minimal effect size among the positive controls, respectively.
 
-The second study was an experiment in which fibroblasts derived from two patients with *SIRT5* mutations were compared to fibroblasts derived from four healthy individuals. 
-The goal of this study was to identify mitochondrial dysfunction in the *SIRT5* mutation carriers. 
-In this study, subjects were nested in group factor (healthy control or patient). 
-Meanwhile, treatment factor (galactose and glucose) and injection factor were crossed with subjects. 
-With both nested and crossed factors, this experiment essentially has a split-plot design. 
-The split plot design was originally invented in agricultural experiments in order to deal randomization of hard-to-change factors. 
-Accordingly, we specified the regression model as:
+The cell line 67375 was measured on two plates. 
+Based on the first assay, our analysis shewed that the posterior probability of patient 67375 having decreasing maximal respiration than the control was 93.0%. 
+By repeating the experiment once more, the posterior probability increased to 99.4%. 
+In contrast, the cell line 76065 was measured on five different plates. 
+Our analysis showed that based on the first assay, the posterior probability of patient 76065 having lower maximal respiration than the control was 40.6%. 
+Repeating the assay once increased the posterior probability from 40.6% to 72.0%. 
+When the assay was repeated for the third, fourth and fifth time, the corresponding posterior probabilities increased to 78.8%, 87.9% and 89.9%, respectively.
 
-$$\mu\text{[i]} = \beta_0 + 
-\beta_G\text{[G[i]]} + 
-\beta_T\text{[T[i]]} + 
-\beta_I\text{[I[i]]} + 
-\beta_{IT}\text{[I[i],T[i]]} + 
-\beta_{GT}\text{[G[i],T[i]]} + 
-\beta_{GI}\text{[G[i],I[i]]} + 
-\beta_S\text{[S[i]]}$$
-
-$\beta_0$ is the overall average OCR per 1000 cells in all wells and across injections. 
-As before, we used $\text{Lognormal(0.54, 0.32)}$ as the prior distribution. 
-$\beta_G$, $\beta_T$ and $\beta_I$ are OCR deflections due to group (healthy or patient), treatment (glucose or galactose), and injections, respectively. 
-To take into account the fact that cell lines derived from the same subject were used for all levels of treatments and injections, $\beta_S$ is used to estimate the deflection due to subject factor. 
-The interaction term, $\beta_{IT}$ denotes whether deflection due to injections is dependent on treatment factor. 
-Similarly, the other two interaction terms, $\beta_{GT}$ and $\beta_{GI}$ denote whether deflection caused by treatment and injection factor are dependent on group factor.
-
-Our analysis showed that on average 1000 fibroblasts treated with galactose had higher basal respiration (1.31 [0.801, 1.86] pmol/min) and ATP-linked respiration (1.17 [0.649, 1.71] pmol/min) than 1000 glucose-treated fibroblasts.
-In contrast, proton leak was similar in galactose or glucose-treated fibroblasts (median difference 0.149 [-0.371, 0.687]). 
-Interestingly, our analysis showed that on average the basal respiration of galactose-treated fibroblasts (1000 cells) was 3.75 [3.33, 4.17] pmol/min in healthy controls and 4.60 [4.07, 5.13] pmol/min in patients with *SIRT5* mutations.
-In glucose-treated fibroblasts, the basal respiration of healthy controls was 2.42 [2.01, 2.84] pmol/min, whereas the basal respiration of patient-derived fibroblasts was 3.29 [2.75, 3.79] pmol/min. 
-The basal respiration difference between patients and controls is 0.861 [0.293, 1.41] and was independent of treatment of glucose or galactose. 
-The ATP-linked respiration in galactose-treated fibroblasts derived from healthy controls and patients with *SIRT5* mutations were 3.35 [2.94, 3.76] and 4.02 [3.47, 4.54], respectively. 
-In glucose-treated fibroblasts, the ATP-linked respiration in healthy controls and patients were 2.17 [1.75, 2.59] pmol/min and 2.85 [2.29, 3.38] pmol/min, respectively. 
-Similar to basal respiration, the difference in ATP-linked respiration between patients and controls was 0.682 [0.0998, 1.24] pmol/min, and this difference was independent of treatment of glucose or galactose.
-The proton leak of galactose-treated fibroblasts in healthy controls was 0.404 [-0.0152, 0.819] pmol/min whereas in patients the proton leak was 0.579 [0.0657, 1.12] pmol/min. 
-In glucose-treated fibroblasts, healthy controls had proton leak 0.258 [-0.164, 0.665] pmol/min, and patients had proton leak 0.434 [-0.0804, 0.962] pmol/min. 
-The proton leak difference between patients and controls was 0.176 [-0.363, 0.736] pmol/min and was independent of treatment of glucose or galactose.
-
-![Figure 2 Posterior distributions of basal respiration, ATP-linked respiration and proton leak of galactose or glucose-treated fibroblasts derived from in healthy subjects and patients with loss-of-function mutations in *SIRT5*](./respiration_sirt5_mutation_study.jpg)
+![Figure 5 (A) Posterior probability of decreasing maximal respiration in patient 67375 compared to the control cell line in the scenario with 1 or 2 plates.
+(B) Posterior probability of decreasing maximal respiration in patient 76065 compared to the control cell line in the scenario with 1, 2, 3, 4 or 5 plates](./posterior_max_resp_67375_76065.jpeg)
 
@@
 
-Although Seahorse XF analyzer is widely used in bioenergetic profiling, its data analysis has not received sufficient attention [@Yepez2018]. 
-In the current study, we developed a Bayesian hierarchical modeling approach, which accounts for the complex structure of Seahorse OCR data. 
-The complexity of Seahorse OCR data is due to the fact that Seahorse XF measurements are nested within injections, injections are crossed with wells, and wells are nested in experimental groups. 
-The data structure can be even more complex when a study has a complicated experimental design with multiple factors.
+Although Seahorse XF analyzer is widely used in bioenergetic profiling, its data analysis has not received sufficient attention. 
+A hidden feature of seahorse XF OCR data is that it contains a complex data structure. 
+The complex data structure is due to the fact that seahorse measurement cycles are nested within injections, injections are crossed with wells, and wells are nested in plates. 
+As far as we know, currently there is no data analysis protocols that take into account this complex data structure, impairing the robustness of seahorse XF OCR data analysis outcomes. 
+This is because when one ignores the data structure, one also ignores the variations between measurement cycles, between wells and between plates. 
+In order to make the seahorse data analysis more robust, in this study we developed a Bayesian hierarchical modeling approach, OCRbayes, which accounts for all these technical variations during the data analysis.
 
-From our perspective, there are three levels including 1) measurement, 2) well and 3) group in the Seahorse XF OCR data. 
-At the lowest level, the measurement level, the variation within an injection phase are assumed to be caused by measurement error. 
-This is because there are multiple measurement cycles during each injection phase, and every measurement cycle starts with a mix and wait steps to ensure the same baseline of cell values, cell physiology should not substantially change within an injection phase.
-Since OCR data must be positive values, we used the lognormal distribution as the likelihood function. 
-The same likelihood function was also used by OCR-Stats [@Yepez2018]. 
-In the first example, we found that on average the uncertainty in FCCP-related measurements (0.147 [0.0443,0.270]) was larger than initial (0.0707 [0.0227,0.176]), oligomycin (0.0923 [0.0257,0.403]) and rotenone/antimycin A (0.0972 [0.0242,0.519]) (Figure S1).
+## Seahorse XF OCR measurements are noisy
 
-![Figure S1 Uncertainty in OCR measurements per scenario in the example 1](./measurement_error_inhibitor_study.jpg)
+An OCR value is determined not only by mitochondrial activity, but also by technical noise including 1) between measurement cycle variation, 2) between well variation and 3) between plate variation.
 
-The variable OCR values could indicate that FCCP possibly induces an unstable physiologic state, possibly due to the off-target effects. 
-It has been proposed that other mitochondrial uncouplers such as BAM15 [@Kenwood2014] would have less off-target effects and would thus generate a more stable cellular physiologic state.
+Every phase contains three measurement cycles, resulting in three OCR values. 
+Since every measurement cycle starts with a "mix and wait" steps to ensure the same baseline of cell values, cell physiology should not substantially change within a phase. 
+Existing tools such as OCR-stats [@Yepez2018] and SHORE [@Nicholas2017] use different strategies to select a single data point to represent an injection phase. 
+When different data points were chosen, one can get different outcomes, making the current seahorse XF data analysis less robust. 
+To avoid this ambiguity, our approach did not select any particular data point, instead modeled all three OCR data. 
+By doing so, we incorporated the uncertainty about "which data point should I choose?" into the data analysis and focused on the average behavior.
 
-In the second example, we found that on average the measurement uncertainty in initial, oligomycin and rotenone/antimycin A were 0.0581 [0.0135, 0.221], 0.0760 [0.0220, 0.262] and 0.0798 [0.0208, 0.301], respectively.
-These measurement uncertainties were comparable between these two case studies, indicating that OCR values were stable in initial, oligomycin and rotenone/antimycin A scenario.
+The between well variation refers to a common observation that OCR values differ among the replicate wells.
+One important factor leading to the variation between replicate wells is that cell numbers are not identical in these wells. 
+The wells with more cells would have higher OCR than the wells containing fewer cells. 
+Fortunately, cell number in each well can be quantified experimentally and used for the data analysis [@Little2020; @Janssen2021]. 
+In addition to cell number difference, initial conditions, treatment concentration, or fluorophore sleeve calibration can also contribute to variation between wells. 
+OCRbayes also takes into account the between well variation caused by these unobserved factors.
 
-![Figure S2 Uncertainty in OCR measurements per scenario in the example 2](./measurement_error_mutation_study.jpg)
+Between plate variation takes place when the same seahorse experiment is repeated on different days and on more than one plates. 
+Due to differences in temperature, seeding time, growth time, growth medium, sensor cartridge as well as treatment efficiency, the OCR outcomes will differ between plates [@Koopman2016; @Yepez2018]. 
+Often the between plate variation is assumed to be the dominant technical variation involved in seahorse OCR data. 
+Based on our analysis, we showed that this assumption may be appropriate for OCR measurements derived from the initial and FCCP phase, but may not work for the OCR values derived from oligomycin and rotenone phase.
 
-The second level of Seahorse OCR data is the well.  
-Since the same cell lines with the same treatment were seeded in multiple wells, these replicated wells (also called pseudo replication) essentially reflect the same biological signals. 
-In other words, these replicate wells should give the same OCR value if the number of cells are identical.
-Treating these replicated wells as biological replicates in t test or ANOVA (analysis of variance) deviates from the assumptions that are associated with those procedures, making these statistical analysis not suitable for experimental data based on a single cell line.
-We notice that a "workaround" is to treat replicated experiments on different days or on different plates as a biological replication. 
-However, this "workaround" does not solve the issue that there is no biological replicate. 
-In fact, this issue is not unique to Seahorse OCR data analysis, it also happens to most if not all cell line-based studies. 
-Our Bayesian hierarchical modeling approach provides a way to properly handle this type of experimental data. 
-However, without biological replicates we still cannot generalize what we observed in one cell line to other cells. 
-Furthermore, since the number of cells are not exactly the same in replicated wells, it is common to see OCR values differ among the pseudo replication [@Little2020]. 
-To adjust for the cell number variation, our hierarchical model explicitly incorporated the cell number information of each wel0l (Equation 7) and estimated the $\text{OCR}_\text{per 1k cells}$, which represented the OCR value of 1000 cells.
+## Comparison with OCR-Stats
 
-The last level of Seahorse OCR data is the experimental group. 
-$\text{OCR}_\text{per 1k cells}$ reflects OCR values that are determined by a particular cell line, a particular treatment and a particular injection. 
-The cell line, treatment and their combination are viewed as groups. 
-To find out the difference between groups, regression analysis of $\text{OCR}_\text{per 1k cells}$ needs to be performed.
-In our case studies, we constructed two different regression models to handle Seahorse XF data with factorial and split-plot designs.
+In this work, we compared the maximal respiration in 26 patient cell lines to the control cell line individually as what was done in OCR-stats [@Yepez2018]. 
+Overall our analysis recalled successfully all negative controls and five out of six positive controls.
 
-Our Bayesian hierarchical modeling approach has several advantages. 
-The first advantage of our approach is that the uncertainty in the OCR measurements was incorporated in the posterior estimation of bioenergetic measures. 
-The second advantage is that our approach is flexible and can be adapted to different experimental designs. 
-The third advantage is that our approach can work with any sample size and unbalanced group sizes. 
-For example, case study 2 has four biological replicates in the control group whereas two in the patient group. 
-This is especially useful for the study that is challenging to acquire biological samples.
+Patient cell line 61818 was labeled as a positive control since this patient was found having a mutation in *SFXN4* gene.
+A recent study based on erythroleukemic cell line showed that *SFXN4* knockout resulted in significant decrease in all parameters of respiration, including baseline respiration, respiratory ATP synthesis, maximal respiration, and spare respiratory capacity [@Paul2019].
+However, our analysis showed that the maximal respiration of this patient was not significantly different from the control cell line. 
+Our further analysis showed that the patient cell line only showed tenancy of having lower maximal respiration than control in one of the three repeated experiments.
+
+We also noticed that our outcomes were not identical as what was presented in OCR-Stats [@Yepez2018]. 
+Strikingly, in our analysis patient 73387 showed significant decreasing in maximal respiration compared to the control cell line. 
+However, in the OCR-stats [@Yepez2018], the difference in maximal respiration was not statistically significant in this patient. 
+This patient 73387 was diagnosed carrying a homozygous loss of function mutation in the *PET100* gene, which encodes a mitochondrial complex IV biogenesis factor [@Yepez2018; @Olahova2015; @Lim2014]. 
+A homozygous truncating variant (c.142C>T, p.(Gln48\*)) in the *PET100* gene was found to lead to a complete loss of enzyme activity, and caused deficiency in complex IV [@Olahova2015a]. 
+We think that the significant reduction of maximal respiration in patient 73387 corresponded to its deficiency in mitochondrial complex IV.
+
+## OCRbayes is potentially used for screening mitochondrial disease
+
+Since increasing in proton leak or a decreasing in basal or maximal respiration are indicators of mitochondrial dysfunction [@Brand2011], seahorse XF is potentially to be used for screening mitochondrial disease patients.
+An immediate question is whether we need to run a single or multiple seahorse assays for a patient. 
+We suggest that whether performing a single or multiple seahorse assay depends on the effect size of perturbation. 
+OCRbayes allows us to calculate posterior probability that the maximal respiration was decreased in patient even based on a single seahorse assay. 
+When the perturbation effect was large, such as cell line 67375, which had the largest effect size among all the positive controls. 
+The posterior probability after observing a single plate was already over 90%. 
+It is reasonable to run just a single assay in this case. 
+In contrast, when the perturbation effect was relatively small, for example the cell line 76065, which had the smallest effect size among all the positive controls, the posterior probability after running a single assay was modest (about 40%).
+Repeating the experiment once more increased the posterior probability from 40% to 70%. 
+Thus in this case, it is beneficial to run multiple seahorse assays. 
+We think that the posterior probability is an useful metric to help scientists to decide whether to perform extra seahorse assays on the patient cell lines.
+
+## Advantages and limitations
+
+OCRbayes has several advantages. 
+The first advantage of our approach is that it incorporates various technical variations including between measurement cycle variation, between well variation as well as between plate variation during the estimation of bioenergetic measures.
+All current methods needs to choose a single data point from the three measurement cycles to represent the OCR during a particular phase. 
+This procedure ignores the uncertainty and makes analysis less robust because different choice of data points may lead to different results. 
+The second advantage is that OCRbayes can calculate posterior probability for difference in various bioenergetic measures based on seahorse OCR data consisting of a single plate or multiple plates. 
+This is a useful feature for screening biopsy samples derived from patients with mitochondrial diseases.
+
+One limitation of this study is that OCRbayes currently focused on human fibroblasts cell line. 
+This is because our model development was restricted to the limited publicly available seahorse OCR data sets. 
+In the future, OCRbayes will be extended to other cell types when more experimental data are generated.
 

### 2021-03-14 manuscript.Rmd
version for biorxiv and npj systems biology and application

@@
 bibliography: /home/xiang/Documents/Bibliography/seahorse.bib
-csl: plos-one.csl
+csl: scientific-reports.csl
 ---
@@
 
-Xiang Zhang, 
-Taolin Yuan, 
-Keijer Jaap,
-Vincent C. J. de Boer
+Xiang Zhang^1,2,\*^, 
+Taolin Yuan^2^, 
+Jaap Keijer^2^,
+Vincent C. J. de Boer^2,\*^
 
-Human and Animal Physiology, Wageningen University, De Elst 1, 6708 WD Wageningen, The Netherlands
+1. Theoretical Biology and Bioinformatics, Utrecht University, Padualaan 8, 3584 CH Utrecht, The Netherlands
+2. Human and Animal Physiology, Wageningen University, De Elst 1, 6708 WD Wageningen, The Netherlands
 
@@
 To rigorously incorporate the complex structure into data analysis, here we developed a Bayesian hierarchical modeling framework, OCRbayes, and demonstrated its applicability based on analysis of published data sets.
-We showed that OCRbayes can analyze seahorse XF OCR data with either single or multiple plates, and it has potential to be used for diagnosing patients with mitochondrial diseases.
+We showed that OCRbayes can analyze Seahorse XF OCR data with either single or multiple plates, and it has potential to be used for diagnosing patients with mitochondrial diseases.
 
@@
 Mitochondria are double-membrane organelles that are central hubs in regulating energy generation and partitioning. 
-Since defects in mitochondrial function have been found to be involved in cardiovascular diseases [@Ballinger2005], type II diabetes [@Szendroedi2011], neurodegenerative disease [@Johri2012] and cancer [@Erez2015], restoring mitochondrial function is emerging to be a therapeutic target for common diseases [@Murphy2018].
+Patients with genetic defects in mitochondrial function are often affected by severe and progressive disease in early life [@Koene2013]. 
+Furthermore, mitochondrial disorders have also been found to be involved in cardiovascular diseases [@Ballinger2005], type II diabetes [@Szendroedi2011], neurodegenerative disease [@Johri2012] and cancer [@Erez2015].
+Thus, restoring mitochondrial function is emerging to be a therapeutic target for both common diseases [@Murphy2018] as well as genetic mitochondrial diseases [@Weissig2020].
 
-Mitochondria produce energy primarily by transferring electrons along the electron transport chain (ETC) on the inner membrane. 
-Along the ETC, there are four complexes (including complex I, II, III and IV), building up a proton gradient that is ultimately used by ATP synthase to generate ATP. 
-Since electrons are primarily accepted by O~2~ to produce H~2~O, the activity of ETC can be assessed by measuring oxygen consumption rate (OCR). 
-As a reliable and efficient platform, Seahorse XF analyzer is a multiwell plate based respirometry assay that is widely used to quantify OCR in living cells [@Wu2007]. 
-Typically, the seahorse XF analyzer measures the OCR of cells in a 96-well plate under different ETC perturbation scenarios, used for assessing mitochondrial functionality such as maximal respiration, leak respiration and ATP-linked respiration [@@Chacko2014].
-Seahorse XF Cell Mito Stress Test consists of four phases including an initial phase, followed by three consecutive injections, including oligomycin, FCCP (carbonyl cyanide-4-(trifluoromethoxy)phenylhydrazone) and a mixture of rotenone and antimycin A.
-In the initial phase, no compound is injected and ETC is not perturbed. 
-Once oligomycin is injected in the assay, proton translocation through ATP synthase is blocked, leading to a reduction in OCR. 
-Next, injection of FCCP allows protons to move into the mitochondrial matrix independent of the ATP synthase, pushing the oxygen consumption to the maximum level. 
-In the end, a mixture of rotenone and antimycin A is injected to inhibit complex I and III, shutting down mitochondrial respiration.
+Mitochondria produce energy in oxidative phosphorylation (OXPHOS) primarily by transferring electrons along the electron transport chain (ETC) on the inner membrane. 
+Along the ETC, there are four complexes (complex I, II, III and IV), together building up a proton gradient that is ultimately used by ATP synthase to generate ATP. 
+Since electrons are primarily accepted by O~2~ to produce H~2~O, the OXPHOS activity can be assessed by measuring oxygen consumption rate (OCR). 
+As a reliable and efficient platform, Seahorse XF analyzer provides a multiwell plate based respirometry assay that is widely used to quantify OCR in living cells [@Wu2007]. 
+Typically, the Seahorse XF analyzer measures the OCR of cells in a 96-well plate under different ETC/OXPHOS perturbation scenarios, used for assessing mitochondrial functionality such as maximal respiration, leak respiration and ATP-linked respiration [@Chacko2014].
 
-A standard seahorse run includes three measurement cycles for each phase. 
-Every measurement cycle starts by lowering the cartridge and creating a temporary semi-closed 7 $\mu$L chamber [@Gerencser2009]. 
+A typical Seahorse run includes three measurement cycles for each phase. 
+Every measurement cycle starts by lowering the cartridge and creating a temporary semi-closed ~2 $\mu$L chamber [@Gerencser2009]. 
 During a measurement cycle, fluorescent oxygen sensors capture oxygen concentration changes in the chamber and outputs OCR. 
-The Seahorse XF analyzer measures OCR in tens to hundreds of thousands of cells per well and typically requires 4–5 replicate wells per experimental group [@Artyomov2020].
-A single 96-well plate can contain multiple experimental groups. 
+The Seahorse XF analyzer measures OCR in tens to hundreds of thousands of cells per well and typically requires 4–5 replicate wells per experimental group, allowing the analysis of multiple experimental groups in one plate [@Artyomov2020].
 Since measurement cycles are nested within phases, phases are crossed with wells, and wells are nested in experimental groups, a very complex structure is inherently embedded in the Seahorse XF OCR data.
 Surprisingly, current Seahorse XF data analyses most often ignore this complex structure, and by default, data sets are often chopped into subgroups followed by performing ANOVA-like statistical tests.
-Although there are tools developed for seahorse data analysis, none of them takes the complexity of the data structure into account. 
-In addition, these methods differ in how to extract OCR value for each phase. 
-For instance, OCR-Stats [@Yepez2018] took OCR value from the last measurement cycle in phase 1, the minimal OCR value in phase 2, the maximal OCR value in phase 3 and the minimal OCR value in phase 4.
-In contrast, SHORE [@Nicholas2017] took the mean OCR value in phase 1, the median OCR value in phase 2, the maximal OCR value in phase 3, and the median OCR value in phase 4.
+Although there are advanced tools developed for Seahorse data analysis [@Nicholas2017;@Yepez2018], none of them comprehensively take the complexity of the data structure into account. 
+This can limit not only the interpretation of Seahorse XF OCR outcomes, but also the development of mitochondrial targeted therapies as well as our understanding of mitochondrial biology.
 
@@
 OCRbayes is based on a fully Bayesian approach due to its flexibility and convenience for constructing the hierarchical models. 
-We analyzed a published OCR data set OCR data set to demonstrate the applicability of our approach analyzing Seahorse XF OCR data as well as its potential usage for mitochondrial disease diagnostics.
+To demonstrate the applicability of our approach as well as its potential implication for mitochondrial disease diagnostics, we applied OCRbayes to a publicly available OCR data set [@Yepez2018], which contains over 200 Seahorse experiments performed on human fibroblasts derived from patients with mitochondrial diseases and controls.
 
@@
 
-To demonstrate how OCRbayes can be used for Seahorse XF OCR data analysis, we analyzed 26 human fibroblasts that were derived from patients that were suspected to have mitochondrial dysfunction. 
-In particular, six of these cell lines were derived from patients with rare mutations in genes associated with an established cellular respiratory defect, and they serve as positive controls. 
-Meanwhile, there are two cell lines that showed no respiratory defects in earlier studies [@Powell2015;@Kremer2016] and they served as negative controls. 
-Since fold change of maximal respiration is the most common metric used for identifying mitochondrial dysfunction, we compared the maximal respiration of all these 26 cell lines against the control cell line.
+## OCRbayes: A Bayesian hierarchical modeling framework for Seahorse OCR data analysis
 
-## Data
+In order to incorporate the complex data structure into the analysis of Seahorse XF OCR measurements, we developed a Bayesian hierarchical modeling framework, OCRbayes. 
+In this study, we focused on experimental data containing two groups, such as patient and control group. 
+From our perspective, the Seahorse OCR data include three levels, including 1) measurement cycle, 2) well and 3) plate.
 
-We downloaded the seahorse XF OCR data presented in OCR-Stats [@Yepez2018] from the journal website https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0199938#sec022. 
-This data set contains 203 human fibroblasts assayed in 126 plates. 
-The normal human dermal fibroblasts (NHDF) is used as control. 
-The rest 202 cell lines were derived from patients suspected to suffer from rare mitochondrial diseases. 
-Among these 202 cell lines, 26 fibroblast cell lines were grown independently and were measured in multiple plates. 
-The rest 176 fibroblast cell lines were measured in a single plate. 
-We then spitted the data set into two subsets. 
-One subset, which contains the 176 fibroblasts that were assayed in a single plate, was used for estimating prior distributions. 
-The other subset, which contains the 26 fibroblasts, was used for benchmarking.
+During a measurement cycle, fluorescent oxygen sensors capture oxygen concentration changes in the chamber and outputs OCR. 
+For every interval, multiple measurement cycles are performed in order to accurately measure the OCR. 
+A typical Seahorse run contains four intervals. 
+The first interval refers to initial phase, and the second, third, fourth interval refer to phase after injecting oligomycin (blocking proton translocation through ATP synthase), FCCP (allowing protons to move into the mitochondrial matrix independent of the ATP synthase) and antimycin/rotenone (inhibiting complex I and III, shutting down mitochondrial respiration).
+The variation between the OCR values within an interval is the between measurement cycle variation. 
+Since OCR values must be positive, we used a lognormal distribution with the true OCR value at the log scale ($\text{logOCR}_\text{true}$) and the between measurement cycle standard deviation (also at the log scale, $\text{logOCR}_\text{sd}$) to model each observed OCR value ($\text{OCR}_\text{obs}$).
 
-Since dealing missing observations is not the focus of this study, we removed the entire measurement values of a well if it contains a missing OCR value. 
-In addition, since we focused on the standard mitochondrial stress test assay, we removed any well if that well does not have 4 phases with 3 measurement cycles each. 
-For the prior distribution subset, we had 177 (including the control cell line NHDF) cell lines on 78 plates after filtering. 
-Since NHDF cell line was present in all 78 plates, the OCR values derived from NHDF cell line were used for estimating the prior distribution of between plate variation.
+$$\text{OCR}_\text{obs}\text{[i]} \sim \text{Lognormal(logOCR}_\text{true}\text{[P[i],W[i],I[i]]}, \text{logOCR}_\text{se}\text{[P[i],I[i]])} \text{ (1)}$$
 
-## Implementation details
+$\text{OCR}_\text{obs}$ is a vector of length $\text{N}_\text{plate} \times \text{N}_\text{well} \times \text{N}_\text{interval} \times \text{N}_\text{measurement}$, where $\text{N}_\text{plate}$, $\text{N}_\text{well}$, $\text{N}_\text{interval}$ and $\text{N}_\text{measurement}$ are number of plates, wells, intervals and measurement cycles, respectively.
+The $\text{logOCR}_\text{true}$ refers to $\text{N}_\text{plate}$ layers of matrices. 
+Each layer is a matrix with $\text{N}_\text{well}$ rows and $\text{N}_\text{interval}$ columns. 
+$\text{logOCR}_\text{sd}$ is a matrix of $\text{N}_\text{plate}$ rows and $\text{N}_\text{interval}$ columns.
 
-The Bayesian hierarchical models were implemented in Stan (version 2.19.3) [@Carpenter2017]. 
+In a typical Seahorse XF assay, the same cell line from the same experimental group is seeded in multiple wells. 
+However, the OCRs in these replicate wells will not be the same, and the difference is called between well variation. 
+One obvious reason causing the between well variation is that the number of cells in these replicate wells are not identical. 
+To adjust for the effect of cell number difference, we modeled the true OCR value as a function of the cell number, and from there estimated $\text{OCR}_\text{per 1k cells}$, which represents OCR value per 1000 cells that received the same treatment (or from the same group) and injection.
+In addition to cell number difference, technical, procedural or instrumental noise can also contribute to between well variation. 
+We captured this well-to-well variation after accounting for cell number difference with the residual parameter $\sigma_\text{well}$.
+
+$$\text{logOCR}_\text{true}\text{[P[i],W[i],I[i]]} \sim \text{Normal(}\text{OCR}_\text{per 1k cells}\text{[P[i],G[i],I[i]]} \times \text{N}_\text{cells}\text{[W[i]]}, \sigma_\text{well}\text{[P[i],G[i],I[i]])} \text{ (2)}$$
+
+$\text{OCR}_\text{per 1k cells}$ is a three-dimensional matrix with $\text{N}_\text{plate}$ layers, and each layer has $\text{N}_\text{group}$ rows and $\text{N}_\text{Interval}$ columns. 
+$\text{N}_\text{cells}$ is a vector of length $\text{N}_\text{well}$, and every entry represents the cell number in that well.
+
+The Seahorse XF assay for a (patient) cell line or a specific condition is often repeated on different days and on more than one plates. 
+Due to batch effects such as plating, culturing or environmental differences between time and laboratories, OCR measurements will differ between plates. 
+To take into account the between plate variation, we used another lognormal distribution with the logarithm transformed OCR value per 1000 cells,  ($\mu_{\text{OCR}_\text{per 1k cells}}$) and the between plate standard deviation ($\sigma_\text{plate}$) to model the OCR value per 1000 cells $\text{OCR}_\text{per 1k cells}$.
+
+$$\text{OCR}_\text{per 1k cells}\text{[P[i],G[i],I[i]]} \sim \text{Lognormal(log(}\mu_{\text{OCR}_\text{per 1k cells}}\text{[G[i], I[i]])}, \sigma_\text{plate}\text{[G[i], I[i]])} \text{ (3)}$$
+
+## Bayesian inference
+
+OCRbayes focuses on calculating posterior distributions for OCR per 1000 cells in various experimental conditions ($\text{P(}\mu_{\text{OCR}_\text{per 1k cells}} \text{|} \text{OCR}_\text{obs}\text{)}$).
+The posterior distributions were combinations of prior distributions ($\text{P(logOCR}_\text{se}\text{)}$, $\text{P(}\sigma_\text{well}\text{)}$, $\text{P(}\sigma_\text{plate}\text{)}$ and $\text{P(}\mu_\text{OCR}_\text{per 1k cells}\text{)}$), and the likelihood function ($\text{P(}\text{OCR}_\text{obs} \text{|} \text{N}_\text{cell}, \mu_{\text{OCR}_\text{per 1k cells}}, \text{logOCR}_\text{sd}, \sigma_\text{well}, \sigma_\text{plate}\text{)}$)
+
+In this study, since our case studies focused on human fibroblast cells, we used informative prior distributions for $\text{logOCR}_\text{sd}$, $\sigma_\text{well}$, $\sigma_\text{plate}$ and $\mu_{\text{OCR}_\text{per 1k cells}}$.
+Since values of all these parameters must be positive, we used four lognormal distributions. 
+We applied maximum likelihood estimation to calculate the lognormal distribution parameters based on the Seahorse XF OCR data in OCR-stats [@Yepez2018] by running the fitdistr function built in the MASS r package.
+In this study, our prior distributions are
+
+$$
+\begin{align}
+\text{logOCR}_\text{sd} & \sim \text{Lognormal(-3.23, 0.79)} \\
+\sigma_\text{well} & \sim \text{Lognormal(-1.62, 0.53)} \\
+\sigma_\text{plate} & \sim \text{Lognormal(-1.18, 0.05)} \\
+\mu_{\text{OCR}_\text{per 1k cells}} & \sim \text{Lognormal(0.3, 0.79)}
+\end{align}
+$$
+
+The Bayesian multi-level model of OCRbayes was implemented in Stan (version 2.19.3) [@Carpenter2017]. 
 We fitted the model by running Hamiltonian Markov Chain Monte Carlo. 
@@
 
-$$\text{Basal respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, rotenone} \text{ (Equation 1)}$$
-$$\text{ATP-linked respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, oligomycin} \text{ (Equation 2)}$$
-$$\text{Proton leak} = \text{OCR}_\text{per 1k cells, oligomycin} - \text{OCR}_\text{per 1k cells, rotenone} \text{ (Equation 3)}$$
-$$\text{Spare respiratory capacity} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, initial} \text{ (Equation 4)}$$
-$$\text{Maximal respiration} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, rotenone} \text{ (Equation 5)}$$
+$$\text{Basal respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, antimycin/rotenone} \text{ (4)}$$
+$$\text{ATP-linked respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, oligomycin} \text{ (5)}$$
+$$\text{Proton leak} = \text{OCR}_\text{per 1k cells, oligomycin} - \text{OCR}_\text{per 1k cells, antimycin/rotenone} \text{ (6)}$$
+$$\text{Spare respiratory capacity} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, initial} \text{ (7)}$$
+$$\text{Maximal respiration} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, antimycin/rotenone} \text{ (8)}$$
+
+## Human fibroblast OCR data
+
+To benchmark OCRbayes as well as illustrate how OCRbayes can be used for analyzing OCR in patients with mitochondrial diseases, we used the OCR data set provided in Yepez et al. [@Yepez2018]. 
+This data set contains Seahorse OCR measurements from 203 human fibroblast cell lines that have been assayed in 126 plates. 
+Normal human dermal fibroblast (NHDF) reference cell lines were used as controls in all plates. 
+The other 202 cell lines were derived from patients suspected to suffer from rare mitochondrial diseases. 
+Among these 202 cell lines, 26 fibroblast cell lines were measured in multiple plates. 
+We used the 176 patient fibroblast cell lines that were assayed in a single plate as well as the control cell line in the same plates for estimating prior distributions. 
+The other 26 fibroblast cell lines were used for benchmarking OCRbayes. 
+Among the 202 cell lines that were analyzed, Yepez et al. [@Yepez2018] labeled 6 patient cell lines as positive controls that have shown statistically significant reduction in maximum respiration. 
+Meanwhile, Yepez et al. also labeled another two patient cell lines as negative controls, since these cell lines did not show changes in OCR in earlier experiments [@Yepez2018].
+
+We processed the original data by removing wells in which single or more OCR measurements were missing. 
+After filtering, we used 176 patient cell lines together with the NHDF control cell line on 78 plates for estimating the prior distributions for between measurement cycle variation ($\text{logOCR}_\text{sd}$) and between well variation ($\sigma_\text{well}$).
+To estimate the prior distribution for between plate variation ($\sigma_\text{plate}$) and mean OCR per 1000 cells ($\mu_{\text{OCR}_\text{per 1k cells}}$), we used the OCR values from the NHDF cell lines that were plated in all 78 plates.
 
@@
 
-For each cell line study, we compared the patient cell line to the control cell line (NHDF). 
-Since maximal respiration is the primary readout of mitochondrial defect in literature, and was also evaluated in the OCR-Stats, here we reported the mean log2 fold change of maximal respiration together with the False Discovery Rate (FDR).
-FDR was calculated based on the posterior error probability ($1 - P(log2(\frac{\text{patient}}{\text{control}}|\text{data}) < 0)$), where $P$ represents the posterior distribution after fitting the Bayesian model to the experimental OCR data.
-If the FDR below 0.05, we consider that difference between patient and control cell line was statistically significant.
+For benchmarking OCRbayes, we compared the patient cell lines to the control cell line (NHDF). 
+Since mitochondrial dysfunction is often reflected in the maximal respiration, which was also evaluated previously in the Yepez et al.
+study [@Yepez2018], we reported here the mean log2 fold change of maximal respiration together with the False Discovery Rate (FDR).
+FDR was calculated based on the posterior error probability ($1 - \text{P(log}_2\text{(}\frac{\text{patient}}{\text{control}}\text{|}\text{OCR}_{obs}) < 0\text{)}$), where $\text{P(log}_2\text{(}\frac{\text{patient}}{\text{control}}\text{|}\text{OCR}_{obs}\text{)}$ represented the posterior distribution of fold change after fitting the Bayesian model to the experimental OCR data.
+If the FDR was below 0.05, we considered that the difference between patient and control cell line was statistically significant.
 
@@
 
-## OCRbayes: A Bayesian hierarchical modeling framework for Seahorse OCR data analysis
+## OCRbayes: from OCR to respiration metric difference
 
-In order to incorporate the complex data structure into the analysis of seahorse XF OCR measurements, we developed a Bayesian hierarchical modeling framework, OCRbayes. 
-In this study, we focused on experimental data containing two groups, patient and control. 
-The mathematical description of our Bayesian hierarchical modeling framework is shown below:
+To demonstrate the applicability, we applied OCRbayes to analyze OCR data derived from two patient cell lines with known mutations in either *BOLA3* or *PET100* gene. 
+These two genes encode proteins that are essential for biogenesis or assembly of mitochondrial complexes [@Olahova2015;@Lim2014;@Haack2013]. 
+We compared the patient cell line to the control cell lines within their respective plates. 
+These two cell lines were both assayed on two plates on two different days. 
+For both cell lines, we observed a clear decrease in maximal respiration compared to the control cell lines in both plates (Figure 1A and 1E). 
+Meanwhile, we noticed that the range of OCR values in the two plates used for profiling *BOLA3* patient cell line differed considerably (Figure 1A). 
+In particular, maximal OCR value in the first plate was around 200 pmol/min, whereas in the second plate the maximal OCR value was around 100 pmol/min. 
+In contrast, the range of OCR values in the two plates used for profiling *PET100* were similar to each other. 
+In addition, it was obvious that there was considerable variation between the replicate wells in both plates for both cell lines.
 
-$$\text{OCR}_\text{obs}\text{[i]} \sim \text{Lognormal(logOCR}_\text{true}\text{[P[i],W[i],I[i]]}, \text{logOCR}_\text{se}\text{[P[i],I[i]])} \text{ (Equation 6)}$$
-$$\text{logOCR}_\text{true}\text{[P[i],W[i],I[i]]} \sim \text{Normal(}\text{OCR}_\text{per 1k cells}\text{[P[i],G[i],I[i]]} \times \text{N}_\text{cells}\text{[W[i]]}, \sigma_\text{well}\text{[P[i],G[i],I[i]])} \text{ (Equation 7)}$$
-$$\text{OCR}_\text{per 1k cells}\text{[P[i],G[i],I[i]]} \sim \text{Lognormal(}\mu_{\text{OCR}_\text{per 1k cells}}\text{[G[i], I[i]]}, \sigma_\text{plate}\text{[G[i], I[i]])} \text{ (Equation 8)}$$
+By applying OCRbayes, we combined the two plates for each cell line, calculated the posterior distributions for OCR per 1000 cells during the four intervals of a typical Seahorse assay, including initial phase without injection (Int1), oligomycin phase (Int2), FCCP phase (Int3) and antimycin/rotenone phase (Int4) (Figure 1B and 1F).
+Next, the calculated posterior OCR per 1000 cells for the four intervals were transformed into the various respiration metrics, including ATP-linked respiration, basal respiration, maximal respiration, proton leak and spare respiratory capacity (Figure 1C and 1G).
+In the last step, we compared the respiration metrics in patient cell line to the control cell line, and calculated the posterior distribution for log2(fold change) (Figure 1D and 1H). 
+We observed that both *BOLA3* and *PET100* patient cell line showed reduction in ATP-linked respiration, basal respiration and maximal respiration, compared to the control cell line. 
+On the other hand, these two patient cell lines showed no difference in proton leak and spare respiratory capacity compared to the control cell line (Figure 1D and 1H).
 
-From our perspective, there are three sources of technical variations including 1) between measurement cycle variation, 2) between well variation and 3) between plate variation in the Seahorse XF OCR data.
+## OCRbayes accounts for various technical variations during Seahorse XF OCR data analysis
 
-The between measurement cycle variation refers to the variability between the three OCR values generated by the three measurement cycles within an injection phase. 
-In Equation 6, we used a lognormal distribution with the true OCR value at the log scale ($\text{logOCR}_\text{true}$) and the between measurement cycle standard deviation (also at the log scale, $\text{logOCR}_\text{se}$) to model each observed OCR value ($\text{OCR}_\text{obs}$).
-$\text{OCR}_\text{obs}$ is a vector of length $\text{N}_\text{plates} \times \text{N}_\text{wells} \times \text{N}_\text{injections} \times \text{N}_\text{Measurement}$, where $\text{N}_\text{plates}$, $\text{N}_\text{wells}$, $\text{N}_\text{injections}$ and $\text{N}_\text{Measurement}$ are number of plates, wells, injections and measurement cycles, respectively.
-The $\text{logOCR}_\text{true}$ refers to $\text{N}_\text{plates}$ layers of matrices. 
-Each layer is a matrix with $\text{N}_\text{wells}$ rows and $\text{N}_\text{injections}$ columns. 
-$\text{logOCR}_\text{se}$ is a matrix of $\text{N}_\text{plates}$ rows and $\text{N}_\text{injections}$ columns.
+OCR measurements generated by the Seahorse XF analyzer were affected by various technical variations, including between measurement cycle variation, between well variation and between plate variation.
+To visualize the technical variations, we used the cell line with a genetic mutation in *NSUN3* gene (patient cell line 76065). 
+This cell line was measured on five different plates, allowing us to visualize all three technical variations.
 
-In a typical seahorse XF assay, the same cell line from the same experimental group is often seeded in multiple wells. 
-However the number of cells in these replicate wells are not identical, thus causing variation in OCR values between these replicate wells. 
-To adjust for the effect of cell number difference, in Equation 7 we modeled the true OCR value as a function of the cell number, and from there estimated $\text{OCR}_\text{per 1k cells}$, which represents OCR value per 1000 cells that received the same treatment (or from the same group) and injection.
-$\text{OCR}_\text{per 1k cells}$ is a three dimensional matrix with $\text{N}_\text{Groups}$ layers, and each layer has $\text{N}_\text{Groups}$ rows and $\text{N}_\text{injections}$ columns. 
-$\text{N}_\text{cells}$ is a vector of length $\text{N}_\text{wells}$, and every entry represents the cell number in that well. 
-In addition to cell number difference, initial conditions, treatment concentration, or fluorophore sleeve calibration can also contribute to between well variation. 
-Since these factors were difficult to be measured experimentally, we can only estimate an overall effect based on the residual parameter $\sigma_\text{well}$. 
-Thus, $\sigma_\text{well}$ represents the well-to-well variation after accounting for cell number difference.
+OCRbayes calculated the between measurement cycle variation for each plate during each interval. 
+For each cell line in each plate, OCRbayes calculated the between well variation during each interval.
+Regarding the between plate variation, OCRbayes estimated it for each cell line during each interval. 
+All the variation values were on the log scale. 
+We observed that for both patient and control cell line in the initial phase (Int1) and FCCP phase (Int3), the between plate variation was larger than the between well variation, which itself was larger than the between measurement cycle variations (Figure 2).
+However, in the oligomycin phase (Int2) and antimycin/rotenone phase (Int4), the between plate variation was not always larger than the other two technical variations (Figure 2). 
+In particular, we observed that the between well variation was larger than the between measurement cycle and the between plate variation in three plates during the antimycin/rotenone phase (Int4) in the patient cell line.
 
-Last but not the least, the seahorse XF assay for a patient cell line is often repeated on different days and on more than one plates. 
-Due to differences in temperature, seeding time, growth time, growth medium, sensor cartridge as well as treatment efficiency, the OCR outcomes will differ between plates. 
-To take into account the between plate variation, in Equation 8 we used another lognormal distribution with the true OCR value per 1000 cells at log scale ($\mu_{\text{OCR}_\text{per 1k cells}}$) and the between plate standard deviation ($\sigma_\text{plate}$) to model the OCR value per 1000 cells $\text{OCR}_\text{per 1k cells}$.
+## Benchmark OCRbayes
 
-In this study, since our case studies focused on human fibroblast cells, we used informative prior distributions for $\text{logOCR}_\text{se}$, $\sigma_\text{well}$, $\sigma_\text{plate}$ and $\mu_{\text{OCR}_\text{per 1k cells}}$.
-Since values of all these parameters must be positive, we used four lognormal distributions. 
-We estimated the lognormal distribution parameters based on the Seahorse XF OCR data in OCR-stats [@Yepez2018] by running the fitdistr function built in the MASS r package. 
-In this study, our prior distributions are
-$$
-\begin{align}
-\text{logOCR}_\text{se} & \sim \text{Lognormal(-3.23, 0.79)} \\
-\sigma_\text{well} & \sim \text{Lognormal(-1.62, 0.53)} \\
-\sigma_\text{plate} & \sim \text{Lognormal(-1.18, 0.05)} \\
-\mu_{\text{OCR}_\text{per 1k cells}} & \sim \text{Lognormal(0.3, 0.79)}
-\end{align}
-$$
+To demonstrate that OCRbayes works properly, we applied it to analyze the published Seahorse XF OCR data set containing 26 patient cell lines as well as a control cell line. 
+This data set contains 6 cell lines were labeled as positive controls and 2 cell lines were labeled as negative controls (as explained in the material and methods section).
 
-## OCRbayes uses OCR measurements as inputs and outputs posterior distributions for respiration metrics
+Based on our analysis, we found 6 patient cell lines that had lower maximal respiration compared to the control cell line with False Discovery Rate (FDR) below 0.05 (Figure 3B). 
+Among the 6 patient cell lines that showed statistically significant reduction in maximal respiration compared to the control cell line, the patient cell line 73387 (mutation in *PET100* gene) showed the largest effect whereas the patient cell line 76065 (mutation in *NUSN3* gene) showed the smallest effect (Figure 3A).
+Five of these 6 patient cell lines were actually the positive controls. 
+Meanwhile, the two negative controls (patient cell lines 73901 and 91410) had FDR above 0.05 in our analysis.
 
-OCRbayes works with the raw OCR data, and estimates the OCR per 1000 cells during initial phase (Int1), oligomycin phase (Int2), FCCP phase (Int3) and rotenone phase (Int4) (Figure 1A and 1B). 
-Afterwards, the posterior OCR per 1000 cells values were transformed into various respiration metrics, such as ATP-linked respiration, basal respiration, maximal respiration, proton leak and spare respiratory capacity (Figure 1C).
-As an example, here we showed patient 65126 and the control cell line (NHDF) on the corresponding plates. 
-The posterior distributions showed the range of values for a respiration parameter. 
-The spread of the posterior distributions represented the uncertainty, which is resulted from technical variations that will be discussed in the next section.
+Our analysis successfully recalled both negative controls and five out of six positive controls. 
+Interestingly, our analysis showed that there was no significant reduction of maximal respiration in the cell line with a genetic mutation in *SFXN4* gene (patient 61818) that was previously labeled as positive control [@Yepez2018].
+This patient cell line was measured on three independent plates on three different days. 
+The range of OCR values of these three plates differed, indicating considerable between plate variation (Figure 4A). 
+The OCR profiles of the patient and control cell line were overlapping in the first plate (Figure 4A). 
+In contrast, the OCR profiles derived from the second and third plate showed OCR decreasing in the patient cell line compared to the control cell line, especially during the FCCP phase (Figure 4A).
+Meanwhile, we observed considerable variation between the replicate wells as well as measurement cycles in these Seahorse assays (Figure 4A).  
+In addition, the between measurement cycle variation in the third plate seemed to be larger than the other two plates (Figure 4A). 
+We analyzed these three repeated experiments separately as well as combined them together. 
+Our separate analysis showed that only the second experiment showed difference in maximal respiration (Figure 4B). 
+When we used OCRbayes to analyze all the three plates together, we found that a tendency that the patient cell line had lower maximal respiration than the control cell line. 
+However, this tendency was not statistically significant.
 
-![Figure 1 (A) Raw OCR values for patient 65126 and control cell line in two plates.
-(B) Posterior distributions for OCR per 1000 cells in patient 65126 and control cell line during initial phase (Int1), oligomycin phase (Int2), FCCP phase (Int3) and rotenone phase (Int4).
-(C) Posterior distributions for respiration metrics in patient 65126 and control cell line.](./posterior_respiration_examples.jpeg)
+## OCRbayes can be used to evaluate the probability that a patient fibroblast cell line has an abnormality in mitochondrial respiration.
 
-## OCRbayes accounts for various technical variations during seahorse XF OCR data analysis
+A feature that makes OCRbayes unique from other methods is that OCRbayes can evaluate what is the probability that a patient has abnormality in his or her fibroblast mitochondrial respiration based on a single Seahorse assay or multiple Seahorse assays.
+To demonstrate this feature, we used two patient cell lines that showed significant reduction in maximal respiration in our analysis. 
+One patient cell line was was the *PET100* gene mutation fibroblast (patient cell line 73387) and the other patient cell line was *NSUN3* gene mutation fibroblast (patient cell line 76065). 
+Among the 6 patient cell lines that showed statistically significant reduction in maximal respiration compared to the control cell line, the *PET100* and *NSUN3* cell line showed the largest and smallest effect, respectively (Figure 3B).
 
-To visualize the technical variations, we used the cell line 76065 as an example. 
-This cell line was measured on five different plates, allowing us to visualize all three technical variations.  
-We observed that in the initial phase (Int1) and FCCP phase (Int3), the between plate variation is dominant source of technical variation, followed by between well and between measurement cycle variations.
-However, in the oligomycin phase (Int2) and rotenone phase (Int4), the between well variation can be as important as the between plate variation.
-
-![Figure 2 Technical variations in seahorse XF OCR data of experiments based on cell line 76065 and control cell line.
-Int1, Int2, Int3 and Int4 represent initial, oligomycin, FCCP and rotenone phase.
-All the variations were on log scale.](./technical_variations_76065.jpeg)
-
-## Benchmark
-
-To demonstrate that OCRbayes works properly, we applied it to analyze a published seahorse XF OCR data set containing 26 patient cell lines as well as a control cell line. 
-In a previous study, 6 cell lines were labeled as positive controls and 2 cell lines were labeled as negative controls. 
-Based on our analysis, we found six patient cell lines that had lower maximal respiration compared to the control cell line with False Discovery Rate (FDR) below 0.05. 
-Five of these six cell lines were actually the positive controls. 
-Meanwhile, the two negative controls had FDR above 0.05.
-
-![Figure 3 Maximal respiration change patient vs.
-control on multiple plates.
-(A) average log2 fold change (y-axis) of maximal respiration of all cell lines repeated across plates (x-axis) and their respective controls, sorted by the False Discovery Rate (FDR).
-Red and blue dots represent positive and negative controls, respectively.
-(B) similar to (A), but depicting FDR.
-Red dashed line represents FDR = 0.05.](./benchmark.jpeg)
-
-Our analysis successfully recalled  all negative controls and five out of six positive controls. 
-Interestingly, our analysis showed that there was no significant reduction of maximal respiration in patient 61818, which was used as a positive control. 
-This patient cell line was measured by seahorse XF OCR assay on three independent plates on three different days. 
-We analyzed these three repeated experiments separately as well as combined them together. 
-We found that only the second experiment showed difference in maximal respiration.
-
-![Figure 4 (A) Raw OCR values for patient cell line 61818 in three repeated plates.
-(B) Posterior distributions of maximal respiration of patient 61818 and the control cell line in the same plate.
-The top three panels represent analysis of three repeated experiments separately.
-The last panel represents the data analysis outcome after combing all three repeated experiments.](./posterior_maximal_respiration_61818.jpeg)
-
-## OCRbayes can facilitate decision making about whether testing on multiple plates is necessary
-
-A frequent question regarding seahorse XF analysis is that whether it is enough to perform the assay on a single plate or do we really need repeat the assay multiple times on different plates. 
-To address this question, we used the two positive control cell lines 67375 and 76065 as examples, since in our analysis they were shown to have the maximal and minimal effect size among the positive controls, respectively.
-
-The cell line 67375 was measured on two plates. 
-Based on the first assay, our analysis shewed that the posterior probability of patient 67375 having decreasing maximal respiration than the control was 93.0%. 
-By repeating the experiment once more, the posterior probability increased to 99.4%. 
-In contrast, the cell line 76065 was measured on five different plates. 
-Our analysis showed that based on the first assay, the posterior probability of patient 76065 having lower maximal respiration than the control was 40.6%. 
-Repeating the assay once increased the posterior probability from 40.6% to 72.0%. 
-When the assay was repeated for the third, fourth and fifth time, the corresponding posterior probabilities increased to 78.8%, 87.9% and 89.9%, respectively.
-
-![Figure 5 (A) Posterior probability of decreasing maximal respiration in patient 67375 compared to the control cell line in the scenario with 1 or 2 plates.
-(B) Posterior probability of decreasing maximal respiration in patient 76065 compared to the control cell line in the scenario with 1, 2, 3, 4 or 5 plates](./posterior_max_resp_67375_76065.jpeg)
+The *PET100* patient cell line was measured on two plates. 
+Based on the first assay, our analysis showed that the posterior probability of this *PET100* mutation carrier having lower maximal respiration than the control was 97.8%. 
+By repeating the experiment once more, the posterior probability increased to 99.8% (Figure 5A). 
+In contrast, the cell line with mutation in *NSUN3* gene (patient cell line 76065) was measured on five different plates. 
+Our analysis showed that based on the first assay, the posterior probability of this *NSUN3* mutation carrier having lower maximal respiration than the control was 41.3%. 
+Repeating the assay once increased the posterior probability from 41.3% to 73.4%. 
+When the assay was repeated for the third, fourth and fifth time, the corresponding posterior probabilities increased to 78.7%, 87.7% and 90.0%, respectively (Figure 5B). 
+In summary, we demonstrate that OCRbayes can be used to evaluate the probability that a patient fibroblast cell line has an abnormality in mitochondrial respiration based on a single Seahorse assay or multiple Seahorse assays.
 
@@
 Although Seahorse XF analyzer is widely used in bioenergetic profiling, its data analysis has not received sufficient attention. 
-A hidden feature of seahorse XF OCR data is that it contains a complex data structure. 
-The complex data structure is due to the fact that seahorse measurement cycles are nested within injections, injections are crossed with wells, and wells are nested in plates. 
-As far as we know, currently there is no data analysis protocols that take into account this complex data structure, impairing the robustness of seahorse XF OCR data analysis outcomes. 
+A hidden feature of Seahorse XF OCR data is that it contains a complex data structure. 
+The complex data structure is due to the fact that measurement cycles are nested within injections, injections are crossed with wells, and wells are nested in plates. 
+As far as we know, currently there is no data analysis protocol that takes into account this complex data structure, impairing the robustness of Seahorse XF OCR data analysis outcomes. 
 This is because when one ignores the data structure, one also ignores the variations between measurement cycles, between wells and between plates. 
-In order to make the seahorse data analysis more robust, in this study we developed a Bayesian hierarchical modeling approach, OCRbayes, which accounts for all these technical variations during the data analysis.
+In order to make the Seahorse data analysis more robust, in this study we developed a Bayesian hierarchical modeling approach, OCRbayes, which accounts for all these technical variations during the data analysis.
 
@@
 
-Every phase contains three measurement cycles, resulting in three OCR values. 
-Since every measurement cycle starts with a "mix and wait" steps to ensure the same baseline of cell values, cell physiology should not substantially change within a phase. 
+Every phase typically contains three measurement cycles, resulting in three OCR values. 
+Since every measurement cycle starts with a "mix and wait" step to ensure the same baseline of cell values, cell physiology should not substantially change within a phase. 
 Existing tools such as OCR-stats [@Yepez2018] and SHORE [@Nicholas2017] use different strategies to select a single data point to represent an injection phase. 
-When different data points were chosen, one can get different outcomes, making the current seahorse XF data analysis less robust. 
+When different data points were chosen, one can get different outcomes, making the current Seahorse XF data analysis less robust. 
 To avoid this ambiguity, our approach did not select any particular data point, instead modeled all three OCR data. 
@@
 
-Between plate variation takes place when the same seahorse experiment is repeated on different days and on more than one plates. 
+Between plate variation takes place when the same Seahorse experiment is repeated on different days and on more than one plate. 
 Due to differences in temperature, seeding time, growth time, growth medium, sensor cartridge as well as treatment efficiency, the OCR outcomes will differ between plates [@Koopman2016; @Yepez2018]. 
-Often the between plate variation is assumed to be the dominant technical variation involved in seahorse OCR data. 
-Based on our analysis, we showed that this assumption may be appropriate for OCR measurements derived from the initial and FCCP phase, but may not work for the OCR values derived from oligomycin and rotenone phase.
+Often the between plate variation is assumed to be the dominant technical variation involved in Seahorse OCR data. 
+Based on our analysis, we showed that this assumption may be appropriate for OCR measurements derived from the initial and FCCP phase, but may not work for the OCR values derived from oligomycin and antimycin/rotenone phase.
+OCR values in the oligomycin and antimycin/rotenone phase were very small and possibly close to the detection limit of the Seahorse XF in a well. 
+Thus, it is more challenging to accurately measure the OCR in these phases.
 
-## Comparison with OCR-Stats
+## Comparison with other statistical tool for Seahorse XF OCR data analysis
 
@@
 Overall our analysis recalled successfully all negative controls and five out of six positive controls. 
-
 Patient cell line 61818 was labeled as a positive control since this patient was found having a mutation in *SFXN4* gene.
@@
 However, our analysis showed that the maximal respiration of this patient was not significantly different from the control cell line. 
-Our further analysis showed that the patient cell line only showed tenancy of having lower maximal respiration than control in one of the three repeated experiments.
+Our further analysis showed that the patient cell line only showed a tendency of having lower maximal respiration than control in one of the three repeated experiments.
 
-We also noticed that our outcomes were not identical as what was presented in OCR-Stats [@Yepez2018]. 
-Strikingly, in our analysis patient 73387 showed significant decreasing in maximal respiration compared to the control cell line. 
+We also noticed that other outcomes were also not identical as what was presented in the OCR-Stats publication [@Yepez2018]. 
+Strikingly, in our analysis *PET100* mutation fibroblast (patient cell line 73387) showed a significant decrease in maximal respiration compared to the control cell line. 
 However, in the OCR-stats [@Yepez2018], the difference in maximal respiration was not statistically significant in this patient. 
-This patient 73387 was diagnosed carrying a homozygous loss of function mutation in the *PET100* gene, which encodes a mitochondrial complex IV biogenesis factor [@Yepez2018; @Olahova2015; @Lim2014]. 
+This patient was diagnosed carrying a homozygous loss of function mutation in the *PET100* gene, which encodes a mitochondrial complex IV biogenesis factor [@Yepez2018; @Olahova2015; @Lim2014]. 
 A homozygous truncating variant (c.142C>T, p.(Gln48\*)) in the *PET100* gene was found to lead to a complete loss of enzyme activity, and caused deficiency in complex IV [@Olahova2015a]. 
-We think that the significant reduction of maximal respiration in patient 73387 corresponded to its deficiency in mitochondrial complex IV.
+Therefore, our observation of significant reduction in maximal respiration in *PET100* mutation fibroblast based on the OCRbayes is in line with the observed loss of enzymatic activity in patients carrying the genetic mutations in *PET100*.
+This difference in analysis outcome highlighted the advantage of OCRbayes which used hierarchical models to incorporate the complex data structure into Seahorse OCR data analysis, helping separate the technical variations from the OCR measurements and identify difference in the biological OCR.
 
-## OCRbayes is potentially used for screening mitochondrial disease
+## OCRbayes is potentially used for screening patients with mitochondrial diseases
 
-Since increasing in proton leak or a decreasing in basal or maximal respiration are indicators of mitochondrial dysfunction [@Brand2011], seahorse XF is potentially to be used for screening mitochondrial disease patients.
-An immediate question is whether we need to run a single or multiple seahorse assays for a patient. 
-We suggest that whether performing a single or multiple seahorse assay depends on the effect size of perturbation. 
-OCRbayes allows us to calculate posterior probability that the maximal respiration was decreased in patient even based on a single seahorse assay. 
-When the perturbation effect was large, such as cell line 67375, which had the largest effect size among all the positive controls. 
-The posterior probability after observing a single plate was already over 90%. 
+Since an increase in proton leak or a decrease in basal or maximal respiration are indicators of mitochondrial dysfunction [@Brand2011], Seahorse XF is potentially to be used for screening mitochondrial disease patients.
+An immediate question is whether we need to run a single or multiple Seahorse assays for a patient. 
+We suggest that whether performing a single or multiple Seahorse assay depends on the effect size of the perturbation. 
+OCRbayes allows us to calculate posterior probability that the maximal respiration (or any other respiration metrics) was abnormal in a patient even based on a single Seahorse assay. 
+For example, the patient cell line with mutation in *PET100* gene showed the largest decrease in maximal respiration in our analysis. 
+The posterior probability for this patient having abnormal mitochondrial respiration after observing a single Seahorse assay was already about 98%. 
 It is reasonable to run just a single assay in this case. 
-In contrast, when the perturbation effect was relatively small, for example the cell line 76065, which had the smallest effect size among all the positive controls, the posterior probability after running a single assay was modest (about 40%).
+In contrast, the cell line with mutation in *NSUN3* gene had the smallest effect size among all the patient cell lines that showed significant reduction in maximal respiration. 
+The posterior probability of this patient having impaired maximum respiration after running a single Seahorse assay was modest (about 40%). 
 Repeating the experiment once more increased the posterior probability from 40% to 70%. 
-Thus in this case, it is beneficial to run multiple seahorse assays. 
-We think that the posterior probability is an useful metric to help scientists to decide whether to perform extra seahorse assays on the patient cell lines.
+Thus in this case, it is beneficial to run multiple Seahorse assays. 
+We think that the posterior probability is a useful metric to help scientists to decide whether to perform extra Seahorse assays on the patient cell lines. 
+The ability of posterior probability calculation by OCRbayes not only allows to make better conclusions about the significance of the effect of the experimental perturbation, it can also prevent from repeating unnecessary, often expensive, experiments.
+Furthermore, when the availability of patient material or the number of target cells (specific isolated immune cell subsets) is limited, OCRbayes is valuable to exploit the limited data and facilitate proper validation the experiments.
 
-## Advantages and limitations
+## Strengths and limitations
 
@@
 The first advantage of our approach is that it incorporates various technical variations including between measurement cycle variation, between well variation as well as between plate variation during the estimation of bioenergetic measures.
-All current methods needs to choose a single data point from the three measurement cycles to represent the OCR during a particular phase. 
+All current methods need to choose a single data point from the three measurement cycles to represent the OCR during a particular phase. 
 This procedure ignores the uncertainty and makes analysis less robust because different choice of data points may lead to different results. 
-The second advantage is that OCRbayes can calculate posterior probability for difference in various bioenergetic measures based on seahorse OCR data consisting of a single plate or multiple plates. 
-This is a useful feature for screening biopsy samples derived from patients with mitochondrial diseases.
+The second advantage is that OCRbayes can calculate posterior probability for difference in various bioenergetic measures based on Seahorse OCR data consisting of a single plate or multiple plates. 
+This is a useful feature for screening samples derived from patients with mitochondrial diseases.
 
-One limitation of this study is that OCRbayes currently focused on human fibroblasts cell line. 
-This is because our model development was restricted to the limited publicly available seahorse OCR data sets. 
-In the future, OCRbayes will be extended to other cell types when more experimental data are generated.
+One limitation of this study is that OCRbayes currently focused on human fibroblasts cell lines. 
+This is because our model development was restricted to the limited publicly available Seahorse OCR data sets. 
+In the future, OCRbayes can be extended to other cell types when more experimental data are collected and combined to establish databases with experimental OCR data.
+
+# Data Availability
+
+The dataset supporting the conclusions of this article is included within the article and its additional file.
+
+# Author contributions
+
+XZ developed the OCRbayes and wrote the manuscript. 
+TY edited the manuscript. 
+JK edited the manuscript.
+VB supervised the project and edited the manuscript. 
+All authors read and approved the final manuscript.
+
+# Competing interests
+
+The authors declare that they have no competing interests
+
+# Figure legends
+
+Figure 1 (A) and (E) are OCR profiles for patient cell lines that have a genetic mutation in either BOLA3 (A) or PET100 (E), compared to control cell lines in two plates.
+(B) and (F) are posterior distributions for OCR per 1000 cells in BOLA3 (B) and PET100 (F) mutated patient cell lines during initial phase (Int1), oligomycin phase (Int2), FCCP phase (Int3) and antimycin/rotenone phase (Int4).
+(C) and (G) are posterior distributions for respiration metrics in BOLA3 (C) and PET100 (G) mutated patient cell lines and control cell lines.
+(D) and (H) are posterior distributions of log2 fold change in the respiration metrics between patient and control cell line.
+
+Figure 2 Technical variations in Seahorse XF OCR data of experiments based on the patient cell line with mutation in *NSUN3* gene and the control cell line.
+Int1, Int2, Int3 and Int4 represent initial, oligomycin, FCCP and antimycin/rotenone phase.
+All the variations were on log scale.
+The dots are the posterior mean and the line segment represent the 95% credible interval.
+
+Figure 3 Maximal respiration change patient vs.
+control on multiple plates.
+(A) average log2 fold change (y-axis) of maximal respiration of all cell lines repeated across plates (x-axis) and their respective controls, sorted by the False Discovery Rate (FDR).
+Red and blue dots represent positive and negative controls, respectively.
+(B) similar to (A), but depicting FDR.
+Red dashed line represents FDR = 0.05.
+
+Figure 4 (A) Raw OCR values for the cell line with genetic mutation in *SFXN4* gene (patient cell line 61818) in three repeated plates.
+(B) Posterior distributions of maximal respiration of the cell line with genetic mutation in *SFXN4* gene (patient cell line 61818) and the control cell line in the same plate.
+(C) Posterior distributions of maximal respiration of the cell line  after combing all three repeated experiments.
+
+Figure 5 (A) Posterior probability of decreasing maximal respiration in patient cell line with mutation in *PET100* gene compared to the control cell line in the scenario with 1 or 2 plates.
+(B) Posterior probability of decreasing maximal respiration in patient cell line with mutation in *NUSN3* gene compared to the control cell line in the scenario with 1, 2, 3, 4 or 5 plates.
 

### 2021-03-14 R/OCRbayes.R
OCRbayes R codes

@@ -1,0 +1,14 @@
+source("./R/customized_functions.R")
+library(rstan)
+
+my_stan_mod <- stan_model("seahorse_ocr_multiple_plates.stan")
+
+ocrbayes <- function(df, iter = 2000, num.chains = 4L, num.threads = 4L, max_treedepth = 10, adapt_delta = 0.95) {
+  dat_list_for_stan <- collect_data_for_stan(df)
+  stan_output <- sampling(my_stan_mod, data = dat_list_for_stan, 
+                          iter = iter, 
+                          chains = num.chains, cores = num.threads, 
+                          control = list(max_treedepth = max_treedepth, adapt_delta = adapt_delta))
+  return(stan_output)
+}
+

### 2021-03-14 R/customized_functions.R

@@ -1,0 +1,214 @@
+library(dplyr)
+library(tidybayes)
+library(broom)
+library(ggplot2)
+library(forcats)
+
+find_host_plate <- function(df, target_cell_line) {
+  df %>% 
+    dplyr::select(plate_id, Cell_line) %>% 
+    distinct() %>% 
+    dplyr::filter(Cell_line %in% target_cell_line) %>% 
+    dplyr::pull(plate_id)
+}
+
+aggregate_measurements <- function(df) {
+  # three measurement cycles within an interval
+  dat.lev1 <- df %>% 
+    dplyr::select(Cell_line, Interval, plate_id, well, cell_n) %>% 
+    distinct()
+  
+  dat.lev1$id1 <- row.names(dat.lev1)
+  
+  return(dat.lev1)
+}
+
+aggregate_wells <- function(df, single_plate = TRUE) {
+  
+  dat.lev2 <- df %>% 
+      dplyr::select(Cell_line, Interval, plate_id) %>% 
+      distinct()
+  
+  dat.lev2$id2 <- row.names(dat.lev2)
+  
+  return(dat.lev2)
+}
+
+aggregate_plates <- function(df) {
+  
+  dat.lev3 <- df %>% 
+      dplyr::select(Cell_line, Interval) %>% 
+      distinct()
+    
+  dat.lev3$id3 <- row.names(dat.lev3)
+  
+  return(dat.lev3)
+}
+
+plate_interval <- function(df) {
+  dat.lev4 <- df %>% 
+    dplyr::select(plate_id, Interval) %>% 
+    distinct()
+  
+  dat.lev4$id4 <- row.names(dat.lev4)
+  
+  return(dat.lev4)
+}
+
+collect_data_for_stan <- function(df) {
+  df <- sort_case_dat(df)
+  dat.lev1 <- aggregate_measurements(df)
+  dat.lev2 <- aggregate_wells(dat.lev1)
+  dat.lev3 <- aggregate_plates(dat.lev2)
+  dat.lev4 <- plate_interval(df)
+  
+  dat.lev0.new <- df %>% 
+    left_join(dat.lev1) %>% 
+    left_join(dat.lev4)
+  
+  dat.lev1.new <- dat.lev1 %>% 
+    dplyr::select(-id1) %>% 
+    left_join(dat.lev2)
+  
+  dat.lev2.new <- dat.lev2 %>% 
+    dplyr::select(-id2) %>% 
+    left_join(dat.lev3)
+  
+  dat_list_for_stan <- list(
[…]

### 2021-03-14 code_for_figure1-3.Rmd
R codes for producing all figures in the manuscript

---
title: "Code for Figure 1 to Figure 3"
author: "Xiang Zhang"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(gridExtra)
source("./R/customized_functions.R")
source("./R/OCRbayes.R")
```

```{r}
yepez.ocr <- read.delim("Yepez_sample_ocr_data.txt")
yepez.meta <- read.delim("Yepez_sample_meta_data.txt")

yepez.ocr %>% 
  dplyr::select(plate_id, well) %>% 
  distinct() %>% 
  count(plate_id) %>% 
  ggplot(aes(plate_id, n)) + 
  geom_point() + 
  labs(y = "Number of wells in the data set") + 
  scale_x_discrete(labels = seq(1, length(unique(yepez.ocr$plate_id)))) + 
  scale_y_continuous(breaks = c(25, 50, 75, 92)) + 
  geom_hline(yintercept = 92, linetype = "dashed", color = "red") + 
  theme_bw()

number_wells_should_have <- 92 * length(unique(yepez.ocr$plate_id))
number_wells_have <- yepez.ocr %>% 
  dplyr::select(plate_id, well) %>% 
  distinct() %>% 
  count(plate_id) %>% 
  pull(n)
sum(number_wells_have)
number_wells_should_have - sum(number_wells_have)

target_fibroblasts <- yepez.meta %>% 
  dplyr::filter(N != 1, 
                Fibroblast.id != "NHDF") %>% 
  dplyr::pull(Fibroblast.id)

target_plates <- find_host_plate(yepez.ocr, target_fibroblasts)

my_dat <- yepez.ocr %>% 
  dplyr::filter(plate_id %in% target_plates)

wells_missing_dat <- my_dat %>% 
  dplyr::filter(is.na(OCR)) %>% 
  dplyr::select(plate_id, well) %>% 
  distinct()

wells_imcomplete_cycles <- my_dat %>% 
  anti_join(wells_missing_dat) %>% 
  group_by(plate_id, well, Interval) %>% 
  summarize(measurement_cycles = n()) %>% 
  dplyr::filter(measurement_cycles != 3) %>% 
  dplyr::select(plate_id, well) %>% 
  distinct()

my_dat_final <- my_dat %>% 
  anti_join(wells_missing_dat) %>% 
  anti_join(wells_imcomplete_cycles) %>% 
  dplyr::filter(Cell_line %in% c("NHDF", target_fibroblasts))

my_controls <- my_dat_final %>% 
  dplyr::filter(Cell_line == "NHDF")

my_cases <- my_dat_final %>% 
  dplyr::filter(Cell_line != "NHDF") %>% 
  group_by(Cell_line) %>% 
  nest() %>% 
  rename(fibroblast_id = Cell_line)
```

```{r}
my_cases_final <- my_cases %>% 
  mutate(control_dat = list(my_controls), 
         mcmc_outcome = map2(data, control_dat, run_ocrbayes_per_case))

#readr::write_rds(my_cases_final, "benchmark_study_stan_outputs.rds")
my_cases_final <- readr::read_rds("benchmark_study_stan_outputs.rds")

my_respiration <- my_cases_final %>% 
  mutate(experimental_setup = map2(data, control_dat, experimental_setup_per_case), 
         respiration = map2(mcmc_outcome, experimental_setup, convert_ocr_to_respiration))

my_FCs <- my_respiration %>% 
  dplyr::select(fibroblast_id, respiration) %>% 
  ungroup() %>% 
  unnest(respiration) %>% 
  spread(Cell_line, val) %>% 
  mutate(logFC = log2(patient / NHDF))

my_stats <- my_FCs %>% 
  dplyr::filter(what == "Maximal respiration") %>% 
  group_by(fibroblast_id, what) %>% 
  summarize(PEP = mean(logFC >= 0), # posterior error probability
            average_logFC = mean(logFC)) %>% 
  ungroup() %>% 
  arrange(PEP)

my_fdrs <- add_fdr(my_stats)

my_stats$FDR <- my_fdrs

my_stats <- my_stats %>% 
  mutate(fibroblast_id = fct_inorder(fibroblast_id))

my_colors <- c("Positive controls" = "red", "Negative controls" = "blue")
fig2 <- ggplot(my_stats, aes(fibroblast_id, FDR)) + 
  geom_point(size = 3) +
  scale_y_log10() + 
  geom_point(data = filter(my_stats, fibroblast_id %in% c("65126", "67375", "76065", "61818", "67333", "73804")), aes(color = "Positive controls"), size = 3) + 
  geom_point(data = filter(my_stats, fibroblast_id %in% c("91410", "73901")), aes(color = "Negative controls"), size = 3) + 
  scale_color_manual("", values = my_colors) + 
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") + 
  labs(x = "Cell line, sorted by False Discovery Rate", 
       y = "False Discovery Rate") + 
  ggtitle("B") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), 
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14), 
        legend.position = c(0.9,0.2), 
        legend.background = element_rect(fill = "transparent"), 
        legend.text = element_text(size = 12))

fig1 <- ggplot(my_stats, aes(fibroblast_id, average_logFC)) + 
  #geom_pointrange(aes(ymin = lower_logFC, ymax = upper_logFC)) + 
  geom_point(size = 3) + 
  geom_point(data = filter(my_stats, fibroblast_id %in% c("65126", "67375", "76065", "61818", "67333", "73804")), aes(color = "Positive controls"), size = 3) + 
  geom_point(data = filter(my_stats, fibroblast_id %in% c("91410", "73901")), aes(color = "Negative controls"), size = 3) + 
  scale_color_manual("", values = my_colors) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") + 
  labs(x = "Cell line, sorted by False Discovery Rate", 
       y = "log2 (maximal respiration fold change) \n Patient / Control") + 
  ggtitle("A") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), 
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14), 
        legend.position = c(0.9,0.2), 
        legend.background = element_rect(fill = "transparent"), 
        legend.text = element_text(size = 12))

grid.arrange(fig1, fig2, ncol = 1)
```

```{r}
sketch_figure_1A <- function(target_dat, my_label, gene_label) {
  ggplot(target_dat, aes(as.factor(time), OCR)) + 
  geom_line(aes(group = well, color = factor(Cell_line, levels = c("NHDF", "patient"), 
                                             labels = c("Control",gene_label)))) + 
  scale_color_brewer("Cell line", palette = "Set2") + 
  facet_wrap(~plate_id, ncol = 1) + 
  labs(x = "Measurement cycle", 
       y = "OCR (pmol/min)", 
       title = my_label) + 
  theme_classic() + 
  theme(legend.position = "bottom", 
        axis.title = element_text(size = 14), 
        axis.text = element_text(size = 10),
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12))
}

sketch_figure_1B <- function(my_cases_final, my_label, which_output, gene_label) {
  my_ocr1k <- spread_draws(my_cases_final$mcmc_outcome[[which_output]], mu_OCR_per_1k[id3])
  my_setup <- tag_experimental_setup(target_dat)
  my_ocr1k %>% 
  mutate(id3 = as.character(id3)) %>% 
  left_join(my_setup) %>% 
  ggplot(aes(mu_OCR_per_1k)) + 
  geom_density(aes(color = factor(Cell_line, levels = c("NHDF", "patient"), 
                                  labels = c("Control", gene_label))), 
               size = 1) + 
  labs(x = "OCR per 1000 cells (pmol/min)", 
       title = my_label) + 
  facet_wrap(~Interval, ncol = 1, scales = "free_y") + 
  scale_color_brewer("Cell line", palette = "Set2") + 
  theme_classic() + 
  theme(legend.position = "bottom", 
        axis.title = element_text(size = 14), 
        axis.text = element_text(size = 10),
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12))
}

sketch_figure_1C <- function(my_respiration, my_label, target_cell_line, gene_label) {
  my_respiration %>% 
  dplyr::select(fibroblast_id, respiration) %>% 
  ungroup() %>% 
  unnest(respiration) %>% 
  dplyr::filter(fibroblast_id %in% target_cell_line) %>% 
  mutate(Cell_line = factor(Cell_line, levels = c("NHDF", "patient"), labels = c("Control", gene_label))) %>%  
  ggplot(aes(val)) + 
  geom_density(aes(color = Cell_line), size = 1) + 
  scale_color_brewer("Cell line", palette = "Set2") + 
  facet_wrap(~what, ncol = 1, scales = "free_y") + 
  theme_classic() + 
  labs(x = "pmol/min", 
       title = my_label) + 
  theme(legend.position = "bottom", 
        legend.title = element_text(size = 12), 
        legend.text = element_text(size = 12), 
        strip.text = element_text(size = 12), 
        axis.text = element_text(size = 10))
}

sketch_figure_1D <- function(my_respiration, my_label, target_cell_line) {
  my_respiration %>% 
  dplyr::select(fibroblast_id, respiration) %>% 
  ungroup() %>% 
  unnest(respiration) %>% 
  dplyr::filter(fibroblast_id == target_cell_line) %>% 
  spread(Cell_line, val) %>% 
  mutate(diff_in_respiration = log2(patient) - log2(NHDF)) %>% 
  ggplot(aes(diff_in_respiration)) + 
  geom_histogram(bins = 100, fill = "grey70") + 
  facet_wrap(~what, ncol = 1, scales = "free") + 
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + 
  theme_classic() + 
  labs(x = "log2(fold change)", 
       y = "Frequency", 
       title = my_label) + 
  theme(legend.position = "bottom", 
        legend.title = element_text(size = 12), 
        legend.text = element_text(size = 12), 
        strip.text = element_text(size = 12), 
        axis.text = element_text(size = 10))
}

```

```{r}
target_cell_line <- "67375"
gene_label <- "BOLA3"
which_output <- which(my_cases_final$fibroblast_id == target_cell_line)
target_dat <- collect_data_per_case(my_cases_final$data[[which_output]], my_cases_final$control_dat[[which_output]])

fig1A <- sketch_figure_1A(target_dat, "A", gene_label = "BOLA3")
fig1B <- sketch_figure_1B(my_cases_final, "B", which_output, gene_label = "BOLA3")
fig1C <- sketch_figure_1C(my_respiration, "C", target_cell_line, gene_label = "BOLA3")
fig1D <- sketch_figure_1D(my_respiration, "D", target_cell_line)

target_cell_line <- "73387"
gene_label <- "PET100"
which_output <- which(my_cases_final$fibroblast_id == target_cell_line)
target_dat <- collect_data_per_case(my_cases_final$data[[which_output]], my_cases_final$control_dat[[which_output]])

fig1E <- sketch_figure_1A(target_dat, "E", gene_label = "PET100")
fig1F <- sketch_figure_1B(my_cases_final, "F", which_output, gene_label = "PET100")
fig1G <- sketch_figure_1C(my_respiration, "G", target_cell_line, gene_label = "PET100")
fig1H <- sketch_figure_1D(my_respiration, "H", target_cell_line)

gridExtra::grid.arrange(fig1A, fig1B, fig1C, fig1D, fig1E, fig1F, fig1G, fig1H, nrow = 2, ncol = 4)
```

# Figure 2

```{r}
fit76065 <- my_cases_final %>% 
  dplyr::filter(fibroblast_id == "76065")

dat.lev0 <- case_dat <- collect_data_per_case(fit76065$data[[1]], fit76065$control_dat[[1]])
dat.lev0 <- sort_case_dat(dat.lev0)
dat.lev1 <- aggregate_measurements(dat.lev0)
dat.lev2 <- aggregate_wells(dat.lev1)
dat.lev3 <- aggregate_plates(dat.lev2)

#sigma_well_76065 <- spread_draws(fit76065$mcmc_outcome[[1]], sigma_well[id2])
sigma_well_76065 <- tidy(fit76065$mcmc_outcome[[1]], "sigma_well", conf.int = TRUE, conf.method = "HPDinterval") %>% 
  separate(term, into = c("what", "id2", "nothing"), sep = "\\[|\\]") %>% 
  mutate(nothing = NULL) %>% 
  left_join(dat.lev2) %>% 
  dplyr::select(-id2, -std.error)
#sigma_well_76065 <- sigma_well_76065 %>% 
#  mutate(id2 = as.character(id2)) %>% 
#  left_join(dat.lev2) %>% 
#  ungroup() %>% 
#  dplyr::select(-id2)

#sigma_cycle_76065 <- spread_draws(fit76065$mcmc_outcome[[1]], sdlog_OCR[id4])
dat.lev4 <- plate_interval(dat.lev0)
sigma_cycle_76065 <- tidy(fit76065$mcmc_outcome[[1]], "sdlog_OCR", conf.int = TRUE, conf.method = "HPDinterval") %>% 
  separate(term, into = c("what", "id4", "nothing"), sep = "\\[|\\]") %>% 
  mutate(nothing = NULL) %>% 
  left_join(dat.lev4) %>% 
  dplyr::select(-id4, -std.error)
#sigma_cycle_76065 <- sigma_cycle_76065 %>% 
#  mutate(id4 = as.character(id4)) %>% 
#  left_join(dat.lev4) %>% 
#  dplyr::select(-id4)

#sigma_plate_76065 <- spread_draws(fit76065$mcmc_outcome[[1]], sigma_plate[id3])
sigma_plate_76065 <- tidy(fit76065$mcmc_outcome[[1]], "sigma_plate", conf.int = TRUE, conf.method = "HPDinterval") %>% 
  separate(term, into = c("what", "id3", "nothing"), sep = "\\[|\\]") %>% 
  mutate(nothing = NULL) %>% 
  left_join(dat.lev3) %>% 
  dplyr::select(-id3, -std.error)
#sigma_plate_76065 <- sigma_plate_76065 %>% 
#  mutate(id3 = as.character(id3)) %>% 
#  left_join(dat.lev3) %>% 
#  ungroup() %>% 
#  dplyr::select(-id3)

cell_int_plate <- sigma_well_76065 %>% 
  dplyr::select(Cell_line:plate_id)

sigma_cycle_76065 <- cell_int_plate %>% 
  left_join(sigma_cycle_76065)

sigma_plate_76065 <- cell_int_plate %>% 
  left_join(sigma_plate_76065)

sigma_well_76065 %>% 
  bind_rows(sigma_plate_76065) %>% 
  bind_rows(sigma_cycle_76065) %>% 
  mutate(what = factor(what, levels = c("sdlog_OCR", "sigma_well", "sigma_plate"), 
                       labels = c("Between measurement cycle", "Between well", "Between plate")), 
         Cell_line = factor(Cell_line, labels = c("Control", "NSUN3"))) %>% 
  ggplot(aes(plate_id, estimate)) + 
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high, color = what), position = position_dodge(width = 0.5, )) + 
  facet_grid(Interval~Cell_line) + 
  scale_color_brewer("", palette = "Set1") + 
  labs(x = "Plate ID", y = "Technical variation (log scale)") + 
  theme_bw() + 
  theme(legend.position = "top", 
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10), 
        axis.title = element_text(size = 12),
        panel.grid = element_blank(), 
        legend.text = element_text(size = 12), 
        strip.text = element_text(size = 12))

```

### 2021-03-14 code_for_figure4.Rmd
R codes for producing all figures in the manuscript

---
title: "Codes for Figure 4"
author: "Xiang Zhang"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(ggplot2)
library(forcats)
library(rstan)
library(tidybayes)
source("./R/customized_functions.R")
source("./R/OCRbayes.R")
```

# Load OCR-stats OCR data

```{r}
yepez.ocr <- read.delim("Yepez_sample_ocr_data.txt")
yepez.meta <- read.delim("Yepez_sample_meta_data.txt")

yepez.df <- yepez.ocr %>% 
  as_tibble()
```

## Plates containing cell line 61818

```{r}
plates_contain_61818 <- find_host_plate(yepez.df, target_cell_line = "61818")

plates_contain_61818
```

There are `r length(plates_contain_76065)` plates that contained cell line 76065.

## Missing data

```{r}
yepez.df %>% 
  dplyr::filter(plate_id %in% plates_contain_61818, 
                is.na(OCR))
```

```{r include=FALSE}
plate_to_remove <- yepez.df %>% 
  dplyr::filter(plate_id %in% plates_contain_61818, is.na(OCR)) %>% 
  pull(plate_id)
```

Here, we can see that there is a missing OCR value for well H5 of plate `r plate_to_remove`. 
Since dealing missing observation is not the focus of this study, I will remove all the data related this well from the analysis.

```{r}
dat61818 <- yepez.df %>% 
  dplyr::filter(plate_id %in% plates_contain_61818, 
                Cell_line %in% c("NHDF", "61818"), 
                !((plate_id == "20130930_1109") & (well == "G3")))
```

## How many wells (per group) are there in each plate

```{r}
dat61818 %>% 
  dplyr::select(plate_id, well, Cell_line) %>% 
  distinct() %>% 
  group_by(plate_id, Cell_line) %>% 
  summarize(number_of_wells = n()) %>% 
  knitr::kable()
```

# First experiment

```{r}
dat.lev0 <- dat61818 %>% 
  dplyr::filter(plate_id %in% c("20130930_1109"))

experiment_setup <- tag_experimental_setup(dat.lev0)

dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_1_plate <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(max_treedepth = 10))

respiration_1_plate <- convert_ocr_to_respiration(scenario_1_plate, experiment_setup)
#respiration_1_plate$scenario <- "1 plate"
respiration_1_plate$plate_id <- "20130930_1109"
```

# Repeat the experiment

```{r}
dat.lev0 <- dat61818 %>% 
  dplyr::filter(plate_id %in% c("20140202_1849"))

experiment_setup <- tag_experimental_setup(dat.lev0)

dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_2_plates <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(max_treedepth = 10))

respiration_2_plates <- convert_ocr_to_respiration(scenario_2_plates, experiment_setup)
respiration_2_plates$plate_id <- "20140202_1849"
```

# Repeat the experiment third time

```{r}
dat.lev0 <- dat61818 %>% 
  dplyr::filter(plate_id %in% c("20140205_1620"))

experiment_setup <- tag_experimental_setup(dat.lev0)

dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_3_plates <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(max_treedepth = 10, adapt_delta = 0.95))

respiration_3_plates <- convert_ocr_to_respiration(scenario_3_plates, experiment_setup)
respiration_3_plates$plate_id <- "20140205_1620"
```

```{r}
dat.lev0 <- dat61818 %>% 
  dplyr::filter(plate_id %in% c("20130930_1109", "20140202_1849", "20140205_1620"))

experiment_setup <- tag_experimental_setup(dat.lev0)

dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_3_plates <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(max_treedepth = 10, adapt_delta = 0.95))

respiration_3_plates_combine <- convert_ocr_to_respiration(scenario_3_plates, experiment_setup)
respiration_3_plates_combine$plate_id <- "3 Plates Combined"
```

```{r}
post_respiration <- respiration_1_plate %>% 
  bind_rows(respiration_2_plates) %>% 
  bind_rows(respiration_3_plates) %>% 
  #bind_rows(respiration_3_plates_combine) %>% 
  dplyr::mutate(plate_id = factor(plate_id))

dat61818 %>% 
  dplyr::filter(plate_id == "20130930_1109", 
                Cell_line == "NHDF", 
                Interval == "Int4") %>% 
  arrange(desc(OCR))

fig1 <- ggplot(dat61818, aes(as.factor(time), OCR)) + 
  geom_line(aes(group = well, color = factor(Cell_line, levels = c("NHDF", "61818"), 
                                             labels = c("Control", "Patient")))) + 
  scale_color_brewer("Cell line", palette = "Set2") + 
  facet_wrap(~plate_id, ncol = 1) + 
  labs(x = "Measurement cycle", y = "OCR (pmol/min)", 
       title = "A") + 
  theme_bw() + 
  theme(legend.position = "bottom", 
        axis.title = element_text(size = 14), 
        axis.text = element_text(size = 10),
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12), 
        panel.grid = element_blank())

fig2 <- post_respiration %>% 
  dplyr::filter(what == "Maximal respiration") %>% 
  mutate(Cell_line = factor(Cell_line, labels = c("Control", "Patient"))) %>%  
  ggplot(aes(val)) + 
  geom_density(aes(color = Cell_line), size = 1) + 
  scale_color_brewer("Cell line", palette = "Set2") + 
  facet_wrap(~plate_id, ncol = 1, scales = "free_y") + 
  labs(x = "Maximal respiration (pmol/min)", 
       title = "B") + 
  theme_bw() + 
  theme(legend.position = "bottom", 
        axis.title = element_text(size = 14), 
        axis.text = element_text(size = 10),
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12), 
        panel.grid = element_blank())

fig3 <- respiration_3_plates_combine %>% 
  dplyr::filter(what == "Maximal respiration") %>% 
  mutate(Cell_line = factor(Cell_line, labels = c("Control", "Patient"))) %>%  
  ggplot(aes(val)) + 
  geom_density(aes(color = Cell_line), size = 1) + 
  scale_color_brewer("Cell line", palette = "Set2") + 
  facet_wrap(~plate_id, ncol = 1, scales = "free_y") + 
  labs(x = "Maximal respiration (pmol/min)", 
       title = "C") + 
  theme_bw() + 
  theme(legend.position = "bottom", 
        axis.title = element_text(size = 14), 
        axis.text = element_text(size = 10),
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 12), 
        panel.grid = element_blank())

gridExtra::grid.arrange(fig1, fig2, fig3, nrow = 1)
```

### 2021-03-14 code_for_figure5.Rmd
R codes for producing all figures in the manuscript

---
title: "Codes for Figure 5"
author: "Xiang Zhang"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(ggplot2)
library(forcats)
library(rstan)
library(tidybayes)
source("./R/customized_functions.R")
source("./R/OCRbayes.R")
```

# Load OCR-stats OCR data

```{r}
yepez.ocr <- read.delim("Yepez_sample_ocr_data.txt")
yepez.meta <- read.delim("Yepez_sample_meta_data.txt")

yepez.df <- yepez.ocr %>% 
  as_tibble()
```

## Plates containing cell line 67375

```{r}
plates_contain_73387 <- find_host_plate(yepez.df, target_cell_line = "73387")

plates_contain_73387
```

There are `r length(plates_contain_67375)` plates that contained cell line 67375.

## Missing data

```{r}
yepez.df %>% 
  dplyr::filter(plate_id %in% plates_contain_73387, 
                is.na(OCR))
```

```{r include=FALSE}
plate_to_remove <- yepez.df %>% 
  dplyr::filter(plate_id %in% plates_contain_73387, is.na(OCR)) %>% 
  dplyr::select(plate_id, well) %>% 
  distinct()
```

Here, we can see that there is a missing OCR value for well H5 of plate `r plate_to_remove`. 
Since dealing missing observation is not the focus of this study, I will remove all the data related this well from the analysis.

```{r}
dat73387 <- yepez.df %>% 
  dplyr::filter(plate_id %in% plates_contain_73387, 
                Cell_line %in% c("73387", "NHDF")) %>% 
  anti_join(plate_to_remove)
```

```{r}
ggplot(dat73387, aes(time, OCR)) + 
  geom_line(aes(group = well, color = Cell_line)) + 
  facet_wrap(~plate_id, ncol = 1)
```

## How many wells (per group) are there in each plate

```{r}
dat73387 %>% 
  dplyr::select(plate_id, well, Cell_line) %>% 
  distinct() %>% 
  group_by(plate_id, Cell_line) %>% 
  summarize(number_of_wells = n()) %>% 
  knitr::kable()
```

# First experiment

```{r}
dat.lev0 <- dat73387 %>% 
  dplyr::filter(plate_id %in% c("20140202_1849"))

experiment_setup <- tag_experimental_setup(dat.lev0)
dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_1_plate <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(adapt_delta = 0.95))

respiration_1_plate <- convert_ocr_to_respiration(scenario_1_plate, experiment_setup)
respiration_1_plate$scenario <- "1 plate"
```

# Repeat the experiment

```{r}
dat.lev0 <- dat73387

experiment_setup <- tag_experimental_setup(dat.lev0)
dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_2_plates <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(adapt_delta = 0.95))

respiration_2_plates <- convert_ocr_to_respiration(scenario_2_plates, experiment_setup)
respiration_2_plates$scenario <- "2 plates"
```

```{r}
post_respiration <- respiration_1_plate %>% 
  bind_rows(respiration_2_plates) %>% 
  dplyr::mutate(scenario = factor(scenario, levels = c("1 plate", 
                                                       "2 plates")))
post_respiration %>% 
  dplyr::filter(what == "Maximal respiration") %>% 
  ggplot(aes(val)) + 
  geom_density(aes(color = Cell_line), size = 1) + 
  scale_color_brewer("", palette = "Set2") + 
  facet_wrap(~scenario, ncol = 1, scales = "free_y") + 
  labs(x = "Maximal respiration") + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

```{r}
fig1 <- post_respiration %>% 
  spread(Cell_line, val) %>% 
  mutate(patient_vs_control = `73387` - NHDF) %>% 
  dplyr::filter(what == "Maximal respiration") %>% 
  group_by(scenario) %>% 
  summarize(prob_lower = mean(patient_vs_control < 0)) %>% 
  ggplot(aes(scenario, prob_lower)) + 
  geom_point(size = 3) + 
  labs(x = "Scenario", y = "Posterior Probability of Decreasing Maximal Respiration", 
       title = "A") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title = element_text(size = 14))
```

## Plates containing cell line 76065

```{r}
plates_contain_76065 <- find_host_plate(yepez.df, target_cell_line = "76065")

plates_contain_76065
```

There are `r length(plates_contain_76065)` plates that contained cell line 76065.

## Missing data

```{r}
yepez.df %>% 
  dplyr::filter(plate_id %in% plates_contain_76065, 
                is.na(OCR))
```

```{r include=FALSE}
plate_to_remove <- yepez.df %>% 
  dplyr::filter(plate_id %in% plates_contain_76065, is.na(OCR)) %>% 
  dplyr::select(plate_id, well) %>% 
  distinct()
```

Here, we can see that there is a missing OCR value for well H5 of plate `r plate_to_remove`. 
Since dealing missing observation is not the focus of this study, I will remove all the data related this well from the analysis.

```{r}
dat76065 <- yepez.df %>% 
  dplyr::filter(plate_id %in% plates_contain_76065, 
                Cell_line %in% c("76065", "NHDF")) %>% 
  anti_join(plate_to_remove)
```

```{r}
ggplot(dat76065, aes(time, OCR)) + 
  geom_line(aes(group = well, color = Cell_line)) + 
  facet_wrap(~plate_id, ncol = 1)
```

## How many wells (per group) are there in each plate

```{r}
dat76065 %>% 
  dplyr::select(plate_id, well, Cell_line) %>% 
  distinct() %>% 
  group_by(plate_id, Cell_line) %>% 
  summarize(number_of_wells = n()) %>% 
  knitr::kable()
```

# First experiment

```{r}
dat.lev0 <- dat76065 %>% 
  dplyr::filter(plate_id %in% c("20131220_911"))

experiment_setup <- tag_experimental_setup(dat.lev0)
dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_1_plate <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(adapt_delta = 0.95))

respiration_1_plate <- convert_ocr_to_respiration(scenario_1_plate, experiment_setup)
respiration_1_plate$scenario <- "1 plate"
```

# Repeat the experiment

```{r}
dat.lev0 <- dat76065 %>% 
  dplyr::filter(plate_id %in% c("20131220_911", "20140115_1314"))

experiment_setup <- tag_experimental_setup(dat.lev0)
dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_2_plates <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(adapt_delta = 0.95))

respiration_2_plates <- convert_ocr_to_respiration(scenario_2_plates, experiment_setup)
respiration_2_plates$scenario <- "2 plates"
```

# Repeat the experiment third time

```{r}
dat.lev0 <- dat76065 %>% 
  dplyr::filter(plate_id %in% c("20131220_911", "20140115_1314", "20140226_1046"))

experiment_setup <- tag_experimental_setup(dat.lev0)
dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_3_plates <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(adapt_delta = 0.95))

respiration_3_plates <- convert_ocr_to_respiration(scenario_3_plates, experiment_setup)
respiration_3_plates$scenario <- "3 plates"
```

# Repeat the experiment fourth time

```{r}
dat.lev0 <- dat76065 %>% 
  dplyr::filter(plate_id %in% c("20131220_911", "20140115_1314", "20140226_1046", "20140226_2147"))

experiment_setup <- tag_experimental_setup(dat.lev0)
dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_4_plates <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(adapt_delta = 0.95))

respiration_4_plates <- convert_ocr_to_respiration(scenario_4_plates, experiment_setup)
respiration_4_plates$scenario <- "4 plates"
```

# Repeat the experiment fifth time

```{r}
dat.lev0 <- dat76065

experiment_setup <- tag_experimental_setup(dat.lev0)
dat_list_for_stan <- collect_data_for_stan(dat.lev0)

scenario_5_plates <- sampling(my_stan_mod, data = dat_list_for_stan, chains = 4L, cores = 4L, control = list(adapt_delta = 0.95))

respiration_5_plates <- convert_ocr_to_respiration(scenario_5_plates, experiment_setup)
respiration_5_plates$scenario <- "5 plates"
```

```{r}
post_respiration <- respiration_1_plate %>% 
  bind_rows(respiration_2_plates) %>% 
  bind_rows(respiration_3_plates) %>% 
  bind_rows(respiration_4_plates) %>% 
  bind_rows(respiration_5_plates) %>% 
  dplyr::mutate(scenario = factor(scenario, levels = c("1 plate", 
                                                       "2 plates", 
                                                       "3 plates", 
                                                       "4 plates", 
                                                       "5 plates")))
post_respiration %>% 
  dplyr::filter(what == "Maximal respiration") %>% 
  ggplot(aes(val)) + 
  geom_density(aes(color = Cell_line), size = 1) + 
  scale_color_brewer("", palette = "Set2") + 
  facet_wrap(~scenario, ncol = 1, scales = "free_y") + 
  labs(x = "Maximal respiration") + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

```{r}
fig2 <- post_respiration %>% 
  spread(Cell_line, val) %>% 
  mutate(patient_vs_control = `76065` - NHDF) %>% 
  dplyr::filter(what == "Maximal respiration") %>% 
  group_by(scenario) %>% 
  summarize(prob_lower = mean(patient_vs_control < 0)) %>% 
  ggplot(aes(scenario, prob_lower)) + 
  geom_point(size = 3) + 
  labs(x = "Scenario", y = "Posterior Probability of Decreasing Maximal Respiration", 
       title = "B") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title = element_text(size = 14))
```

```{r}
gridExtra::grid.arrange(fig1, fig2, nrow = 1)
```

### 2021-03-14 code_for_prior_distribution_estimation.Rmd
prior distribution estimation

---
title: "Code for prior distribution estimation"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
source("./R/customized_functions.R")
```

# Load OCR-stats OCR data

```{r}
yepez.ocr <- read.delim("Yepez_sample_ocr_data.txt")
yepez.meta <- read.delim("Yepez_sample_meta_data.txt")
```

# Prior distributions

In the OCR-stats paper, there are in total 203 fibroblasts, in which 176 fibroblasts were measured in a single plate and all these cases have no known genetic defects. 
I will use these cases for estimating prior distributions.

```{r}
orphan_experiments <- yepez.meta %>% 
  dplyr::filter(N == 1) %>% 
  dplyr::pull(Fibroblast.id)

# find the corresponding plates
my_plates <- find_host_plate(yepez.ocr, target_cell_line = orphan_experiments) %>% 
  unique()
```

Next, find the data of the corresponding plates and keep the NHDF and target cell lines data

```{r}
my_dat <- yepez.ocr %>% 
  dplyr::filter(plate_id %in% my_plates, 
                Cell_line %in% c("NHDF", orphan_experiments))

my_dat
```

Procedures:
1. Values of three repeated measurement cycles need to be aggregated into one value
2. Replicate wells need to be aggregated and cell numbers need to be accounted
3. Difference in OCR per 1K cell between a orphan cell line and NHDF
4. Perform 1 - 3 in all 78 plates, separately

```{r}
# remove wells containing NA
wells_contain_NA <- my_dat %>% 
  dplyr::filter(is.na(OCR)) %>% 
  dplyr::select(plate_id, well) %>% 
  unique()

my_dat_filtered <- my_dat %>% 
  anti_join(wells_contain_NA)

wells_imcomplete_cycles <- my_dat_filtered %>% 
  group_by(plate_id, well, Interval) %>% 
  summarize(measurement_cycles = n()) %>% 
  dplyr::filter(measurement_cycles != 3) %>% 
  dplyr::select(plate_id, well) %>% 
  distinct()

my_dat_final <- my_dat_filtered %>% 
  anti_join(wells_imcomplete_cycles)
```

```{r}
lv0 <- my_dat_final

lv1 <- lv0 %>% 
  group_by(plate_id, Cell_line, well, cell_n, Interval) %>% 
  nest() %>% 
  mutate(lnmod = map(data, ~true_signal_measurement_error(.x$OCR))) %>% 
  tidyr::unnest(lnmod) %>% 
  dplyr::select(-data, -std.error) %>% 
  spread(term, estimate) %>% 
  ungroup()
```

First prior distribution I need is for sdlog.

```{r}
ggplot(lv1, aes(sdlog)) + 
  geom_density(aes(color = Interval)) + 
  scale_color_brewer(palette = "Dark2")

MASS::fitdistr(lv1$sdlog, densfun = "lognormal")
```

```{r}
lv2 <- lv1 %>% 
  dplyr::select(-sdlog) %>% 
  mutate(ocr_per_1k = exp(meanlog)/cell_n * 1000) %>% 
  group_by(plate_id, Cell_line, Interval) %>% 
  summarize(ocr_per_1k_true = mean(ocr_per_1k), 
            ocr_per_1k_error_log = sd(log(ocr_per_1k)))
```

The second prior distribution I need is for sigma_well

```{r}
MASS::fitdistr(lv2$ocr_per_1k_error_log, densfun = "lognormal")
```

```{r}
lv3 <- lv2 %>% 
  dplyr::filter(Cell_line == "NHDF") %>% 
  group_by(Cell_line, Interval) %>% 
  summarise(mu_ocr_per_1k_true = mean(ocr_per_1k_true), 
            sd_ocr_per_1k_true_log = sd(log(ocr_per_1k_true)))

lv3
```

The third prior distribution I need is for sigma_plate

```{r}
MASS::fitdistr(lv3$sd_ocr_per_1k_true_log, densfun = "lognormal")
```

The fourth prior distribution I need is OCR_per_1k[1] and OCR_per_1k[5]

```{r}
nhdf_basal <- lv2 %>% 
  dplyr::select(-ocr_per_1k_error_log) %>% 
  dplyr::filter(Cell_line == "NHDF", Interval == "Int1") %>% 
  spread(Interval, ocr_per_1k_true) %>% 
  dplyr::pull(Int1)

nhdf_basal

MASS::fitdistr(nhdf_basal, densfun = "lognormal")

patient_basal <- lv2 %>% 
  dplyr::select(-ocr_per_1k_error_log) %>% 
  dplyr::filter(Cell_line != "NHDF", Interval == "Int1") %>% 
  spread(Interval, ocr_per_1k_true) %>% 
  dplyr::pull(Int1)

patient_basal

MASS::fitdistr(patient_basal, densfun = "lognormal")
```

The fifth prior distribution I need is for OCR_per_1k[2] and OCR_per_1k[6]

```{r}
nhdf_oligomycin <- lv2 %>% 
  dplyr::select(-ocr_per_1k_error_log) %>% 
  dplyr::filter(Cell_line == "NHDF", Interval == "Int2") %>% 
  spread(Interval, ocr_per_1k_true) %>% 
  dplyr::pull(Int2)

nhdf_oligomycin

MASS::fitdistr(nhdf_oligomycin, densfun = "lognormal")

patient_oligomycin <- lv2 %>% 
  dplyr::select(-ocr_per_1k_error_log) %>% 
  dplyr::filter(Cell_line != "NHDF", Interval == "Int2") %>% 
  spread(Interval, ocr_per_1k_true) %>% 
  dplyr::pull(Int2)

patient_oligomycin

MASS::fitdistr(patient_oligomycin, densfun = "lognormal")
```

The sixth prior distribution I need is for OCR_per_1k[3] and OCR_per_1k[7]

```{r}
nhdf_fccp <- lv2 %>% 
  dplyr::filter(Cell_line == "NHDF", Interval == "Int3") %>% 
  dplyr::select(-ocr_per_1k_error_log) %>% 
  spread(Interval, ocr_per_1k_true) %>% 
  dplyr::pull(Int3)

nhdf_fccp

MASS::fitdistr(nhdf_fccp, densfun = "lognormal")

patient_fccp <- lv2 %>% 
  dplyr::filter(Cell_line != "NHDF", Interval == "Int3") %>% 
  dplyr::select(-ocr_per_1k_error_log) %>% 
  spread(Interval, ocr_per_1k_true) %>% 
  dplyr::pull(Int3)

patient_fccp

MASS::fitdistr(patient_fccp, densfun = "lognormal")
```

The sixth prior distribution I need is for OCR_per_1k[4] and OCR_per_1k[8]

```{r}
nhdf_rotenone <- lv2 %>% 
  dplyr::filter(Cell_line == "NHDF", Interval == "Int4") %>% 
  dplyr::select(-ocr_per_1k_error_log) %>% 
  spread(Interval, ocr_per_1k_true) %>% 
  dplyr::pull(Int4)

nhdf_rotenone

MASS::fitdistr(nhdf_rotenone, densfun = "lognormal")

patient_rotenone <- lv2 %>% 
  dplyr::filter(Cell_line != "NHDF", Interval == "Int4") %>% 
  dplyr::select(-ocr_per_1k_error_log) %>% 
  spread(Interval, ocr_per_1k_true) %>% 
  dplyr::pull(Int4)

patient_rotenone

MASS::fitdistr(patient_rotenone, densfun = "lognormal")
```

### 2021-03-16 code_for_figure1-3.Rmd
add codes for calculating posterior mean and 95% credile interval for log2(fold change) of respiration metrics

@@
 
+my_respiration %>% 
+  dplyr::select(fibroblast_id, respiration) %>% 
+  ungroup() %>% 
+  unnest(respiration) %>% 
+  dplyr::filter(fibroblast_id == target_cell_line) %>% 
+  spread(Cell_line, val) %>% 
+  mutate(diff_in_respiration = log2(patient) - log2(NHDF)) %>% group_by(what) %>% summarize(post_mean = mean(diff_in_respiration), low_mean = quantile(diff_in_respiration, 0.025), high_mean = quantile(diff_in_respiration, 0.975))
+
 target_cell_line <- "73387"
@@
 fig1H <- sketch_figure_1D(my_respiration, "H", target_cell_line)
+
+my_respiration %>% 
+  dplyr::select(fibroblast_id, respiration) %>% 
+  ungroup() %>% 
+  unnest(respiration) %>% 
+  dplyr::filter(fibroblast_id == target_cell_line) %>% 
+  spread(Cell_line, val) %>% 
+  mutate(diff_in_respiration = log2(patient) - log2(NHDF)) %>% group_by(what) %>% summarize(post_mean = mean(diff_in_respiration), low_mean = quantile(diff_in_respiration, 0.025), high_mean = quantile(diff_in_respiration, 0.975))
 

### 2021-03-16 code_for_figure4.Rmd
add codes for calculating posterior mean and 95% credible intervals for log2(fold change)

@@
 respiration_1_plate$plate_id <- "20130930_1109"
+
+respiration_1_plate %>% 
+  spread(Cell_line, val) %>% 
+  dplyr::mutate(diff_in_respir = log2(`61818`) - log2(NHDF)) %>% 
+  dplyr::group_by(what) %>% 
+  summarize(post_mean = mean(diff_in_respir), 
+            low_mean = quantile(diff_in_respir, probs = 0.025), 
+            high_mean = quantile(diff_in_respir, probs = 0.975))
 ```
@@
 respiration_2_plates$plate_id <- "20140202_1849"
+
+respiration_2_plates %>% 
+  spread(Cell_line, val) %>% 
+  dplyr::mutate(diff_in_respir = log2(`61818`) - log2(NHDF)) %>% 
+  dplyr::group_by(what) %>% 
+  summarize(post_mean = mean(diff_in_respir), 
+            low_mean = quantile(diff_in_respir, probs = 0.025), 
+            high_mean = quantile(diff_in_respir, probs = 0.975))
 ```
@@
 respiration_3_plates$plate_id <- "20140205_1620"
+
+respiration_3_plates %>% 
+  spread(Cell_line, val) %>% 
+  dplyr::mutate(diff_in_respir = log2(`61818`) - log2(NHDF)) %>% 
+  dplyr::group_by(what) %>% 
+  summarize(post_mean = mean(diff_in_respir), 
+            low_mean = quantile(diff_in_respir, probs = 0.025), 
+            high_mean = quantile(diff_in_respir, probs = 0.975))
 ```
@@
 respiration_3_plates_combine$plate_id <- "3 Plates Combined"
+
+respiration_3_plates_combine %>% 
+  spread(Cell_line, val) %>% 
+  dplyr::mutate(diff_in_respir = log2(`61818`) - log2(NHDF)) %>% 
+  dplyr::group_by(what) %>% 
+  summarize(post_mean = mean(diff_in_respir), 
+            low_mean = quantile(diff_in_respir, probs = 0.025), 
+            high_mean = quantile(diff_in_respir, probs = 0.975))
 ```

### 2021-03-16 manuscript.Rmd
version submited to npj Systems Biology and Application, as well as bioRxiv

@@
 
-Xiang Zhang^1,2,\*^, 
+Xiang Zhang^1,2^, 
 Taolin Yuan^2^, 
@@
 To rigorously incorporate the complex structure into data analysis, here we developed a Bayesian hierarchical modeling framework, OCRbayes, and demonstrated its applicability based on analysis of published data sets.
-We showed that OCRbayes can analyze Seahorse XF OCR data with either single or multiple plates, and it has potential to be used for diagnosing patients with mitochondrial diseases.
+We showed that OCRbayes can analyze Seahorse XF OCR experimental data derived from either single or multiple plates. 
+Moreover, OCRbayes has potential to be used for diagnosing patients with mitochondrial diseases.
 
@@
 
-A typical Seahorse run includes three measurement cycles for each phase. 
+A typical Seahorse assay includes three measurement cycles for each phase. 
 Every measurement cycle starts by lowering the cartridge and creating a temporary semi-closed ~2 $\mu$L chamber [@Gerencser2009]. 
@@
 Although there are advanced tools developed for Seahorse data analysis [@Nicholas2017;@Yepez2018], none of them comprehensively take the complexity of the data structure into account. 
-This can limit not only the interpretation of Seahorse XF OCR outcomes, but also the development of mitochondrial targeted therapies as well as our understanding of mitochondrial biology.
+As a result, OCR variation between measurement cycles, replicate wells and replicate plates is overlooked, impairing the robustness of the interpretation of Seahorse XF OCR outcomes, and eventually the development of mitochondrial targeted therapies as well as our understanding of mitochondrial biology.
 
@@
 For every interval, multiple measurement cycles are performed in order to accurately measure the OCR. 
-A typical Seahorse run contains four intervals. 
+A typical Seahorse assay contains four intervals. 
 The first interval refers to initial phase, and the second, third, fourth interval refer to phase after injecting oligomycin (blocking proton translocation through ATP synthase), FCCP (allowing protons to move into the mitochondrial matrix independent of the ATP synthase) and antimycin/rotenone (inhibiting complex I and III, shutting down mitochondrial respiration).
@@
 
-In a typical Seahorse XF assay, the same cell line from the same experimental group is seeded in multiple wells. 
+In a typical Seahorse XF assay, one cell line undergoing the same experimental treatment is seeded in multiple wells. 
 However, the OCRs in these replicate wells will not be the same, and the difference is called between well variation. 
@@
 
-The Seahorse XF assay for a (patient) cell line or a specific condition is often repeated on different days and on more than one plates. 
+Apart from the technical replicates in one plate, the biological insight for a (patient) cell line or a specific condition is generally validated by repeating the Seahorse assay on different days. 
+As a result, OCR data are distributed on more than one plate. 
 Due to batch effects such as plating, culturing or environmental differences between time and laboratories, OCR measurements will differ between plates. 
@@
 In the last step, we compared the respiration metrics in patient cell line to the control cell line, and calculated the posterior distribution for log2(fold change) (Figure 1D and 1H). 
-We observed that both *BOLA3* and *PET100* patient cell line showed reduction in ATP-linked respiration, basal respiration and maximal respiration, compared to the control cell line. 
-On the other hand, these two patient cell lines showed no difference in proton leak and spare respiratory capacity compared to the control cell line (Figure 1D and 1H).
+We observed that *BOLA3* patient cell line showed reduction in basal respiration (posterior mean log2(fold change) -0.974, 95% credible interval [-1.93, -0.0970]) and maximal respiration (-1.06 [-1.87,-0.228]), compared to the control cell line (Figure 1D).
+On the other hand, we observed no difference in ATP-linked respiration (-1.08 [-2.38, 0.117]), proton leak (-0.630 [-3.87, 2.32]) and spare respiratory capacity (-1.26 [-4.41,1.62]) between *BOLA3* patient cell line and the control cell line (Figure 1D).
+Meanwhile we observed lower ATP-basal respiration (-1.75 [-3.18, -0.546]), basal respiration (-1.58 [-2.52, -0.667]) and maximal respiration (-1.28 [-2.10, -0.436]) in the *PET100* patient cell line than the control cell line (Figure 1H).
+We observed no difference in proton leak (-0.906 [-4.14,2.54]) and spare respiratory capacity (-0.826 [-3.29, 2.34]) in the *PET100* patient cell line compared to the control cell line (Figure 1H).
 
@@
 
-To demonstrate that OCRbayes works properly, we applied it to analyze the published Seahorse XF OCR data set containing 26 patient cell lines as well as a control cell line. 
-This data set contains 6 cell lines were labeled as positive controls and 2 cell lines were labeled as negative controls (as explained in the material and methods section).
+To demonstrate that OCRbayes works properly, we applied it to analyze the published Seahorse XF OCR data set containing 26 patient cell lines as well as a control cell line reported by Yepez et al.
+[@Yepez2018].
+This data set contains 6 cell lines that were labeled as positive controls and 2 cell lines that were labeled as negative controls (as explained in the material and methods section).
 
-Based on our analysis, we found 6 patient cell lines that had lower maximal respiration compared to the control cell line with False Discovery Rate (FDR) below 0.05 (Figure 3B). 
+Based on our analysis, we found 6 patient cell lines that had lower maximal respiration compared to the control cell line with False Discovery Rate (FDR) below 0.05 (Figure 3A and B). 
 Among the 6 patient cell lines that showed statistically significant reduction in maximal respiration compared to the control cell line, the patient cell line 73387 (mutation in *PET100* gene) showed the largest effect whereas the patient cell line 76065 (mutation in *NUSN3* gene) showed the smallest effect (Figure 3A).
-Five of these 6 patient cell lines were actually the positive controls. 
+Five of these 6 patient cell lines were labeled as positive controls in the original study [@Yepez2018]. 
 Meanwhile, the two negative controls (patient cell lines 73901 and 91410) had FDR above 0.05 in our analysis.
@@
 We analyzed these three repeated experiments separately as well as combined them together. 
-Our separate analysis showed that only the second experiment showed difference in maximal respiration (Figure 4B). 
-When we used OCRbayes to analyze all the three plates together, we found that a tendency that the patient cell line had lower maximal respiration than the control cell line. 
-However, this tendency was not statistically significant.
+Our separate analysis showed that none of the three experiments showed significant reduction in maximal respiration in the *SFXN4* patient cell line (Figure 4B). 
+The posterior mean log2(fold change) and the corresponding 95% credible intervals derived from the first, second and third Seahorse assay were 0.122 [-0.834, 1.04], -0.431 [-1.40, 0.499] and -0.0977 [-1.04, 0.827], respectively.
+When we used OCRbayes to analyze all the three plates together, we found that the posterior mean log2(fold change) of maximal respiration was -0.200 [-0.877, 0.500] compared to the control cell line (Figure 4C).
+However, this tendency of reduction in maximal respiration in *SFXN4* patient cell line was not statistically significant.
 
@@
 To demonstrate this feature, we used two patient cell lines that showed significant reduction in maximal respiration in our analysis. 
-One patient cell line was was the *PET100* gene mutation fibroblast (patient cell line 73387) and the other patient cell line was *NSUN3* gene mutation fibroblast (patient cell line 76065). 
+One patient cell line was the *PET100* gene mutation fibroblast (patient cell line 73387) and the other patient cell line was *NSUN3* gene mutation fibroblast (patient cell line 76065). 
 Among the 6 patient cell lines that showed statistically significant reduction in maximal respiration compared to the control cell line, the *PET100* and *NSUN3* cell line showed the largest and smallest effect, respectively (Figure 3B).
@@
 Since an increase in proton leak or a decrease in basal or maximal respiration are indicators of mitochondrial dysfunction [@Brand2011], Seahorse XF is potentially to be used for screening mitochondrial disease patients.
-An immediate question is whether we need to run a single or multiple Seahorse assays for a patient. 
-We suggest that whether performing a single or multiple Seahorse assay depends on the effect size of the perturbation. 
-OCRbayes allows us to calculate posterior probability that the maximal respiration (or any other respiration metrics) was abnormal in a patient even based on a single Seahorse assay. 
+Besides providing solely statistical significance information as other methods do, OCRbayes allows us to calculate posterior probability that the maximal respiration (or any other respiration metrics) was abnormal in a patient even based on a single Seahorse assay.
+This feature is also helpful for deciding whether we need to run a single or multiple Seahorse assays for a patient.
+
 For example, the patient cell line with mutation in *PET100* gene showed the largest decrease in maximal respiration in our analysis. 
@@
 XZ developed the OCRbayes and wrote the manuscript. 
-TY edited the manuscript. 
-JK edited the manuscript.
+TY and JK contributed to the conceptualization of the project and edited the manuscript. 
 VB supervised the project and edited the manuscript. 

### 2021-03-26 manuscript.Rmd
version for BMC bioinformatics

@@
 bibliography: /home/xiang/Documents/Bibliography/seahorse.bib
-csl: scientific-reports.csl
+csl: bmc-bioinformatics.csl
 ---
@@
 
-Mitochondrial dysfunction is involved in many complex diseases. 
+Background: Mitochondrial dysfunction is involved in many complex diseases. 
 Efficient and accurate evaluation of mitochondrial functionality is crucial for understanding pathology as well as facilitating novel therapeutic developments. 
@@
 Surprisingly, statistical analysis of Seahorse XF data has not received sufficient attention, and current methods completely ignore the complex data structure, impairing the robustness of statistical inference.
-To rigorously incorporate the complex structure into data analysis, here we developed a Bayesian hierarchical modeling framework, OCRbayes, and demonstrated its applicability based on analysis of published data sets.
-We showed that OCRbayes can analyze Seahorse XF OCR experimental data derived from either single or multiple plates. 
+
+Results: To rigorously incorporate the complex structure into data analysis, here we developed a Bayesian hierarchical modeling framework, OCRbayes, and demonstrated its applicability based on analysis of published data sets.
+
+Conclusions: We showed that OCRbayes can analyze Seahorse XF OCR experimental data derived from either single or multiple plates. 
 Moreover, OCRbayes has potential to be used for diagnosing patients with mitochondrial diseases.
+
+Keywords: Oxygen consumption rate; mitochondria; Bayesian; hierarchical modeling; seahorse extracellular flux
 
@@
 
-# Data Availability
+## Declarations
 
-The dataset supporting the conclusions of this article is included within the article and its additional file.
+# Ethics approval and consent to participate
 
-# Author contributions
+Not applicable
+
+## Consent for publication
+
+## Availability of data and materials
+
+The datasets analysed during the current study are available in the Github repository, https://github.com/XiangZhangSC/seahorse.
+
+## Competing interests
+
+The authors declare that they have no competing interests
+
+## Funding
+
+Not applicable
+
+## Author's contributions
 
@@
 
-# Competing interests
+## Acknowledgements
 
-The authors declare that they have no competing interests
+Not applicable
 
